---
title: "Stacked Audience–Curator Selection Model"
subtitle: "Paper 1 analysis"
date: "2026-08-13"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
---

# Purpose

This analysis compares audience and editorial/community-curator preferences on
the same outcome scale. Each eligible root comment appears twice: once in an
audience choice set and once in a curator choice set. In an article with (k)
eligible Editors' Picks, curators select the (k) pinned roots and the audience
selects the (k) roots with the highest final relative-vote totals.

A conditional logistic model stratified by article and selector compares
comments only within the same discussion and selection process.
Selector-by-feature interactions directly estimate curator–audience
differences. Standard errors are clustered by article because both selector
choice sets and all comments from a discussion remain dependent.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
suppressPackageStartupMessages({
  library(data.table)
  library(survival)
})
```

# Configuration

The environment variables `COMMENTGAP_TIE_DRAWS` and
`COMMENTGAP_PROBABILITY_DRAWS` control the number of tie sensitivity fits and
probability-interval simulations. Their defaults keep an interactive knit
quick; the submission analysis uses 10 tie realizations and 5,000 probability
draws.

```{r configuration}

input_file <- "data/comment_data_general_091022_untilarticle5874_final_redacted.csv"
output_dir <- "model_output/stacked_selection"
base_seed <- 20260813L
tie_draws <- as.integer(Sys.getenv("COMMENTGAP_TIE_DRAWS", "1"))
probability_draws <- as.integer(Sys.getenv("COMMENTGAP_PROBABILITY_DRAWS", "5000"))

if (is.na(tie_draws) || tie_draws < 1L) {
  stop("COMMENTGAP_TIE_DRAWS must be a positive integer")
}
if (is.na(probability_draws) || probability_draws < 100L) {
  stop("COMMENTGAP_PROBABILITY_DRAWS must be at least 100")
}
if (!file.exists(input_file)) stop("Missing input file: ", input_file)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

required_columns <- c(
  "comment_id", "article", "is_root_comment", "pinned_f",
  "votes_rel", "sentiment_prob_pos", "sentiment_prob_neg",
  "lexdiv_cttr", "SMOG_readability", "cosine_1", "num_words",
  "num_punct", "second_person", "user_follower", "hours_since_article"
)
```

# Data and eligible choice sets

Only root comments in discussions containing at least one root-level Editors'
Pick enter the comparison. Non-root pins are recorded as anomalies and removed
so that curator and audience selections use identical eligible sets.

```{r load-and-filter-data}

cat("Reading analysis columns from", input_file, "\n")
comments <- fread(input_file, select = required_columns)
missing_columns <- setdiff(required_columns, names(comments))
if (length(missing_columns) > 0L) {
  stop("Missing required columns: ", paste(missing_columns, collapse = ", "))
}

if (anyDuplicated(comments$comment_id)) stop("comment_id is not unique")
if (anyNA(comments$article) || anyNA(comments$comment_id)) {
  stop("article and comment_id must not be missing")
}

comments[, pinned_f := as.logical(pinned_f)]
comments[, is_root_comment := as.logical(is_root_comment)]
comments[, second_person := as.integer(as.logical(second_person))]

nonroot_pins <- comments[pinned_f == TRUE & is_root_comment == FALSE, .N]
all_root_comments <- comments[is_root_comment == TRUE]
pin_articles <- all_root_comments[pinned_f == TRUE, unique(article)]
roots <- all_root_comments[article %chin% pin_articles]
roots[, `:=`(
  n_candidates = .N,
  n_picks = sum(pinned_f)
), by = article]

# Conditional choice strata require at least one selected and one unselected
# candidate. One observed discussion selects every eligible root and therefore
# contains no information about relative selection.
uninformative_articles <- roots[n_picks >= n_candidates, unique(article)]
roots <- roots[n_picks > 0L & n_picks < n_candidates]

if (anyNA(roots[, ..required_columns])) {
  na_counts <- vapply(roots[, ..required_columns], anyNA, logical(1))
  stop(
    "Missing values in model inputs: ",
    paste(names(na_counts)[na_counts], collapse = ", ")
  )
}
```

# Predictor transformations

Predictors are limited to comment-intrinsic or approximately posting-time
available attributes. Final votes, pin status, final discussion size, and
downstream reply-tree outcomes are not used as predictors. Log word count is
included to distinguish lexical-diversity and SMOG associations from comment
length.

```{r transform-predictors}

safe_z <- function(x, name) {
  x <- as.numeric(x)
  s <- sd(x)
  if (!is.finite(s) || s <= 0) stop("Cannot standardize ", name)
  (x - mean(x)) / s
}

# Posting-time-available or comment-intrinsic predictors only. Final votes,
# pin status, final discussion size, and reply-tree outcomes are deliberately
# excluded as predictors. log(num_words) adjusts the textual measures for the
# strong association between comment length, lexical diversity, and SMOG.
roots[, z_positive_sentiment := safe_z(sentiment_prob_pos, "sentiment_prob_pos")]
roots[, z_negative_sentiment := safe_z(sentiment_prob_neg, "sentiment_prob_neg")]
roots[, z_lexical_diversity := safe_z(lexdiv_cttr, "lexdiv_cttr")]
roots[, z_reading_difficulty := safe_z(SMOG_readability, "SMOG_readability")]
roots[, z_topical_similarity := safe_z(cosine_1, "cosine_1")]
roots[, z_log_words := safe_z(log1p(num_words), "log1p(num_words)")]
roots[, z_log_punctuation := safe_z(log1p(num_punct), "log1p(num_punct)")]
roots[, z_log_followers := safe_z(log1p(user_follower), "log1p(user_follower)")]
roots[, z_log_posting_hours := safe_z(log1p(hours_since_article), "log1p(hours_since_article)")]
```

```{r feature-metadata}

feature_info <- data.table(
  term = c(
    "z_positive_sentiment", "z_negative_sentiment",
    "z_lexical_diversity", "z_reading_difficulty",
    "z_topical_similarity", "z_log_words", "z_log_punctuation",
    "second_person", "z_log_followers", "z_log_posting_hours"
  ),
  feature = c(
    "Positive sentiment", "Negative sentiment", "Lexical diversity",
    "Reading difficulty (SMOG)", "Topical similarity", "Comment length",
    "Punctuation", "Second-person pronoun", "Author followers",
    "Posting time after article"
  ),
  unit = c(rep("One standard deviation", 7L), "Absent to present",
           rep("One standard deviation", 2L)),
  transformation = c(
    "standardized", "standardized", "standardized", "standardized",
    "standardized", "log1p then standardized", "log1p then standardized",
    "binary", "log1p then standardized", "log1p then standardized"
  )
)
feature_terms <- feature_info$term

scaling_table <- rbindlist(list(
  data.table(raw_variable = "sentiment_prob_pos", transformed_term = "z_positive_sentiment", mean = mean(roots$sentiment_prob_pos), sd = sd(roots$sentiment_prob_pos)),
  data.table(raw_variable = "sentiment_prob_neg", transformed_term = "z_negative_sentiment", mean = mean(roots$sentiment_prob_neg), sd = sd(roots$sentiment_prob_neg)),
  data.table(raw_variable = "lexdiv_cttr", transformed_term = "z_lexical_diversity", mean = mean(roots$lexdiv_cttr), sd = sd(roots$lexdiv_cttr)),
  data.table(raw_variable = "SMOG_readability", transformed_term = "z_reading_difficulty", mean = mean(roots$SMOG_readability), sd = sd(roots$SMOG_readability)),
  data.table(raw_variable = "cosine_1", transformed_term = "z_topical_similarity", mean = mean(roots$cosine_1), sd = sd(roots$cosine_1)),
  data.table(raw_variable = "log1p(num_words)", transformed_term = "z_log_words", mean = mean(log1p(roots$num_words)), sd = sd(log1p(roots$num_words))),
  data.table(raw_variable = "log1p(num_punct)", transformed_term = "z_log_punctuation", mean = mean(log1p(roots$num_punct)), sd = sd(log1p(roots$num_punct))),
  data.table(raw_variable = "log1p(user_follower)", transformed_term = "z_log_followers", mean = mean(log1p(roots$user_follower)), sd = sd(log1p(roots$user_follower))),
  data.table(raw_variable = "log1p(hours_since_article)", transformed_term = "z_log_posting_hours", mean = mean(log1p(roots$hours_since_article)), sd = sd(log1p(roots$hours_since_article)))
))
```

# Construct matched selections

Within vote totals tied at the audience top-(k) boundary, a seeded random key
selects the required number of comments. Sorting by comment ID before drawing
the key makes each realization independent of input row order. The sensitivity
section repeats the model under additional seeds.

```{r selection-functions}

build_audience_selection <- function(root_data, seed) {
  out <- copy(root_data)

  # A seeded random key is used only within equal-vote groups. Sorting by
  # comment_id before drawing makes the realization invariant to input order.
  setorder(out, article, comment_id)
  set.seed(seed)
  out[, tie_key := runif(.N)]
  setorder(out, article, -votes_rel, tie_key)
  out[, audience_selected := seq_len(.N) <= first(n_picks), by = article]
  setorder(out, article, comment_id)

  if (any(out[, sum(audience_selected) != first(n_picks), by = article]$V1)) {
    stop("Audience top-k construction did not select exactly k comments")
  }
  out
}

tie_diagnostics <- roots[, {
  cutoff <- sort(votes_rel, decreasing = TRUE)[first(n_picks)]
  above <- sum(votes_rel > cutoff)
  tied <- sum(votes_rel == cutoff)
  slots <- first(n_picks) - above
  list(
    n_candidates = first(n_candidates),
    n_picks = first(n_picks),
    cutoff_votes_rel = cutoff,
    n_above_cutoff = above,
    n_tied_at_cutoff = tied,
    slots_within_cutoff_tie = slots,
    cutoff_inclusion_probability = slots / tied,
    ambiguous_cutoff = tied > slots
  )
}, by = article]

build_stacked_data <- function(root_data, seed) {
  selected_roots <- build_audience_selection(root_data, seed)
  keep <- c("comment_id", "article", "n_candidates", "n_picks", feature_terms)

  audience <- selected_roots[, ..keep]
  audience[, `:=`(
    selector = "audience",
    curator = 0L,
    selected = as.integer(selected_roots$audience_selected)
  )]

  curator <- selected_roots[, ..keep]
  curator[, `:=`(
    selector = "curator",
    curator = 1L,
    selected = as.integer(selected_roots$pinned_f)
  )]

  stacked <- rbindlist(list(audience, curator), use.names = TRUE)
  stacked[, stratum := interaction(article, selector, drop = TRUE)]
  setorder(stacked, article, selector, comment_id)

  stratum_checks <- stacked[, .(
    selected_n = sum(selected),
    expected_n = first(n_picks),
    candidates_n = .N
  ), by = .(article, selector)]
  if (stratum_checks[, any(selected_n != expected_n)]) {
    stop("At least one selector stratum does not contain exactly k selections")
  }
  if (stratum_checks[, any(selected_n <= 0L | selected_n >= candidates_n)]) {
    stop("At least one selector stratum lacks outcome variation")
  }
  stacked
}
```

# Conditional logistic model

The audience coefficients describe associations with audience top-(k)
selection. Each `curator:feature` interaction is the curator-minus-audience
difference. Article-by-selector strata absorb discussion-specific selection
rates and all other characteristics shared within each choice set.

```{r fit-primary-model}

model_rhs <- c(
  feature_terms,
  paste0("curator:", feature_terms),
  "strata(stratum)",
  "cluster(article)"
)
model_formula <- as.formula(paste("selected ~", paste(model_rhs, collapse = " + ")))

fit_one_draw <- function(seed, keep_data = FALSE) {
  stacked <- build_stacked_data(roots, seed)
  cat("Fitting tie seed", seed, "on", nrow(stacked), "stacked rows\n")
  fit <- clogit(
    model_formula,
    data = stacked,
    method = "efron",
    model = FALSE,
    x = FALSE,
    y = FALSE
  )
  if (anyNA(coef(fit))) stop("Model contains unidentified coefficients")
  if (keep_data) list(fit = fit, data = stacked) else fit
}

primary <- fit_one_draw(base_seed, keep_data = TRUE)
primary_model <- primary$fit
stacked_data <- primary$data
rm(primary)
```

# Selector-specific associations

```{r summarize-associations}

coefficient_table <- data.table(
  term = names(coef(primary_model)),
  estimate = unname(coef(primary_model)),
  std_error_article_clustered = sqrt(diag(vcov(primary_model)))
)
coefficient_table[, `:=`(
  conf_low = estimate - qnorm(0.975) * std_error_article_clustered,
  conf_high = estimate + qnorm(0.975) * std_error_article_clustered,
  z = estimate / std_error_article_clustered
)]
coefficient_table[, p_value := 2 * pnorm(-abs(z))]

summarise_feature <- function(term, label, unit) {
  beta_name <- term
  # R canonicalizes an interaction written curator:x as x:curator.
  difference_name <- paste0(term, ":curator")
  coefficient_names <- names(coef(primary_model))
  if (!(beta_name %in% coefficient_names) || !(difference_name %in% coefficient_names)) {
    stop("Expected model term is absent: ", term)
  }

  b <- coef(primary_model)[beta_name]
  g <- coef(primary_model)[difference_name]
  v <- vcov(primary_model)
  audience_se <- sqrt(v[beta_name, beta_name])
  difference_se <- sqrt(v[difference_name, difference_name])
  curator_se <- sqrt(
    v[beta_name, beta_name] + v[difference_name, difference_name] +
      2 * v[beta_name, difference_name]
  )

  interval <- function(est, se) c(est - qnorm(0.975) * se, est + qnorm(0.975) * se)
  audience_ci <- interval(b, audience_se)
  curator_ci <- interval(b + g, curator_se)
  difference_ci <- interval(g, difference_se)

  data.table(
    term = term,
    feature = label,
    unit = unit,
    audience_log_odds = unname(b),
    audience_se = audience_se,
    audience_conf_low = audience_ci[1],
    audience_conf_high = audience_ci[2],
    audience_odds_ratio = exp(b),
    curator_log_odds = unname(b + g),
    curator_se = curator_se,
    curator_conf_low = curator_ci[1],
    curator_conf_high = curator_ci[2],
    curator_odds_ratio = exp(b + g),
    curator_minus_audience_log_odds = unname(g),
    difference_se = difference_se,
    difference_conf_low = difference_ci[1],
    difference_conf_high = difference_ci[2],
    difference_p_value = 2 * pnorm(-abs(g / difference_se))
  )
}

selector_associations <- rbindlist(Map(
  summarise_feature,
  feature_info$term,
  feature_info$feature,
  feature_info$unit
))
selector_associations[, difference_q_value_bh := p.adjust(
  difference_p_value,
  method = "BH"
)]
```

# Probability-scale reference contrasts

For interpretation, each selector-specific slope is translated into the change
from an article's observed selection prevalence, (p_0=k/N), after a one-unit
feature increase. Continuous predictors are standardized, so their unit is one
standard deviation. These are equal-discussion-weighted reference contrasts,
not claims about causal effects or individual latent preferences.

```{r probability-contrasts}

# Translate the selector-specific slopes to a common, transparent probability
# reference. For each discussion, p0=k/N is its observed selection prevalence.
# The reported contrast is the equal-discussion-weighted change from p0 after a
# one-unit feature increase. Continuous features are standardized, so this is a
# one-SD contrast. Confidence intervals simulate the joint pair of audience and
# curator-difference coefficients using the article-clustered covariance.
article_rates <- unique(roots[, .(article, n_candidates, n_picks)])
article_rates[, p0 := n_picks / n_candidates]

probability_change <- function(beta, p0) {
  mean(plogis(qlogis(p0) + beta) - p0)
}

probability_summary <- function(term, label, unit, draws = probability_draws) {
  beta_name <- term
  difference_name <- paste0(term, ":curator")
  estimates <- coef(primary_model)[c(beta_name, difference_name)]
  covariance <- vcov(primary_model)[c(beta_name, difference_name), c(beta_name, difference_name), drop = FALSE]

  set.seed(base_seed + match(term, feature_terms) * 100L)
  simulated <- MASS::mvrnorm(draws, mu = estimates, Sigma = covariance)

  audience_point <- probability_change(estimates[1], article_rates$p0)
  curator_point <- probability_change(sum(estimates), article_rates$p0)
  difference_point <- curator_point - audience_point

  audience_draws <- apply(simulated, 1L, function(x) probability_change(x[1], article_rates$p0))
  curator_draws <- apply(simulated, 1L, function(x) probability_change(sum(x), article_rates$p0))
  difference_draws <- curator_draws - audience_draws
  interval <- function(x) unname(quantile(x, c(0.025, 0.975), na.rm = TRUE))

  audience_ci <- interval(audience_draws)
  curator_ci <- interval(curator_draws)
  difference_ci <- interval(difference_draws)

  data.table(
    term = term,
    feature = label,
    unit = unit,
    reference_probability_mean = mean(article_rates$p0),
    audience_probability_change = audience_point,
    audience_conf_low = audience_ci[1],
    audience_conf_high = audience_ci[2],
    curator_probability_change = curator_point,
    curator_conf_low = curator_ci[1],
    curator_conf_high = curator_ci[2],
    curator_minus_audience_change = difference_point,
    difference_conf_low = difference_ci[1],
    difference_conf_high = difference_ci[2]
  )
}

probability_contrasts <- rbindlist(Map(
  probability_summary,
  feature_info$term,
  feature_info$feature,
  feature_info$unit
))
probability_contrasts[, `:=`(
  audience_change_pp = 100 * audience_probability_change,
  curator_change_pp = 100 * curator_probability_change,
  curator_minus_audience_pp = 100 * curator_minus_audience_change,
  difference_conf_low_pp = 100 * difference_conf_low,
  difference_conf_high_pp = 100 * difference_conf_high
)]
```

# Vote-tie sensitivity

```{r tie-sensitivity}

tie_sensitivity <- data.table()
tie_sensitivity_summary <- data.table()
if (tie_draws > 1L) {
  tie_results <- vector("list", tie_draws)
  for (draw in seq_len(tie_draws)) {
    seed <- base_seed + draw - 1L
    fit <- if (draw == 1L) primary_model else fit_one_draw(seed)
    tie_results[[draw]] <- data.table(
      draw = draw,
      seed = seed,
      term = feature_terms,
      curator_minus_audience_log_odds = unname(coef(fit)[paste0(feature_terms, ":curator")])
    )
    if (draw > 1L) rm(fit)
  }
  tie_sensitivity <- rbindlist(tie_results)
  tie_sensitivity_summary <- tie_sensitivity[, .(
    draws = .N,
    estimate_median = median(curator_minus_audience_log_odds),
    estimate_min = min(curator_minus_audience_log_odds),
    estimate_max = max(curator_minus_audience_log_odds),
    estimate_sd = sd(curator_minus_audience_log_odds),
    sign_changes_across_draws = min(curator_minus_audience_log_odds) < 0 &
      max(curator_minus_audience_log_odds) > 0
  ), by = term]
  tie_sensitivity_summary <- merge(
    feature_info[, .(term, feature)],
    tie_sensitivity_summary,
    by = "term",
    all.y = TRUE,
    sort = FALSE
  )
}
```

# Sample accounting and exports

```{r sample-summary}

sample_summary <- data.table(
  statistic = c(
    "all_comments", "all_articles", "all_root_comments",
    "articles_with_root_pins", "eligible_root_comments",
    "eligible_articles", "eligible_root_pins", "nonroot_pins_excluded",
    "uninformative_all_roots_selected_articles", "stacked_rows",
    "ambiguous_audience_top_k_cutoff_articles", "mean_article_selection_rate"
  ),
  value = c(
    nrow(comments), uniqueN(comments$article), nrow(all_root_comments),
    length(pin_articles), nrow(roots), uniqueN(roots$article), sum(roots$pinned_f),
    nonroot_pins, length(uninformative_articles), nrow(stacked_data),
    tie_diagnostics[ambiguous_cutoff == TRUE, .N], mean(article_rates$p0)
  )
)
```

```{r export-results, include=FALSE}

fwrite(feature_info, file.path(output_dir, "feature_definitions.csv"))
fwrite(scaling_table, file.path(output_dir, "feature_scaling.csv"))
fwrite(tie_diagnostics, file.path(output_dir, "audience_top_k_tie_diagnostics.csv"))
fwrite(sample_summary, file.path(output_dir, "sample_summary.csv"))
fwrite(coefficient_table, file.path(output_dir, "model_coefficients.csv"))
fwrite(selector_associations, file.path(output_dir, "selector_associations.csv"))
fwrite(probability_contrasts, file.path(output_dir, "probability_contrasts.csv"))
if (nrow(tie_sensitivity) > 0L) {
  fwrite(tie_sensitivity, file.path(output_dir, "tie_sensitivity_draws.csv"))
  fwrite(tie_sensitivity_summary, file.path(output_dir, "tie_sensitivity_summary.csv"))
}

# Strip the call environment before serialization so the model artifact does
# not capture the half-million-row stacked working dataset.
environment(primary_model$terms) <- baseenv()
saveRDS(primary_model, file.path(output_dir, "stacked_selection_model.rds"))

diagnostic_lines <- c(
  "Stacked audience/curator selection model",
  paste("Input:", input_file),
  paste("Primary tie seed:", base_seed),
  paste("Tie realizations fitted:", tie_draws),
  paste("Eligible discussions:", uniqueN(roots$article)),
  paste("Eligible root comments:", nrow(roots)),
  paste("Stacked rows:", nrow(stacked_data)),
  paste("Article-selector strata:", uniqueN(stacked_data$stratum)),
  paste("Model concordance:", unname(summary(primary_model)$concordance[1])),
  paste("Robust score test p-value:", summary(primary_model)$robscore[3]),
  "Audience selection: top k final relative-vote roots, seeded tie-breaking.",
  "Curator selection: k pinned roots from the identical candidate set.",
  "Inference: conditional logit with article x selector strata; SEs clustered by article.",
  "Probability contrasts: equal-discussion-weighted reference changes based on p0=k/N.",
  "Interpretation is associational: final votes are affected by exposure and existing ranking."
)
writeLines(diagnostic_lines, file.path(output_dir, "diagnostics.txt"))
```

# Results

The main inferential quantity is the curator-minus-audience interaction.
Positive values indicate that a characteristic is more strongly associated
with curator selection; negative values indicate a stronger audience
association. Statistical significance alone should not determine substantive
interpretation, so the probability-scale contrasts are reported alongside the
log-odds estimates.

```{r display-results, echo=FALSE}

interaction_display <- selector_associations[, .(
  feature,
  `Curator − audience log-odds` = curator_minus_audience_log_odds,
  `95% CI low` = difference_conf_low,
  `95% CI high` = difference_conf_high,
  `BH q-value` = difference_q_value_bh
)]

probability_display <- probability_contrasts[, .(
  feature,
  `Audience change (pp)` = audience_change_pp,
  `Curator change (pp)` = curator_change_pp,
  `Curator − audience (pp)` = curator_minus_audience_pp,
  `95% CI low (pp)` = difference_conf_low_pp,
  `95% CI high (pp)` = difference_conf_high_pp
)]

knitr::kable(sample_summary, caption = "Analysis sample accounting", digits = 3)
knitr::kable(
  interaction_display,
  caption = "Curator–audience differences on the common selection scale",
  digits = 4
)
knitr::kable(
  probability_display,
  caption = "Equal-discussion-weighted reference probability contrasts",
  digits = 3
)

if (nrow(tie_sensitivity_summary) > 0L) {
  knitr::kable(
    tie_sensitivity_summary,
    caption = "Sensitivity across seeded top-k tie realizations",
    digits = 4
  )
}
```

# Interpretation limits

The audience outcome is based on final relative-vote totals, not a direct
measure of latent preference. Voting can be influenced by posting time,
position, existing ranking, and curator pinning. Likewise, the model estimates
associations within curator-active discussions and should not be generalized
to all discussions or interpreted causally. The `curator` label includes
journalists, community-management staff, and volunteer community guides.
