---
title: "Development-only collinearity sensitivity for stacked selection models"
output:
  html_document:
    toc: true
    toc_depth: 3
---

# Purpose

This analysis diagnoses dependence among terms in the stacked conditional-logit
models and tests prespecified reduced specifications. It is deliberately limited
to the 50% development partition. The shared choice-set parquet contains both
partitions, so test rows are present during initial file loading, but they are
immediately removed by an inner join to development article IDs. No Paper 2 test
row enters scaling, fitting, scoring, diagnostics, or model comparison, and no
sealed-test performance artifact is read.

The primary full models in `06B_stacked_selection_models.Rmd` remain the fixed
reference. Reduced models are robustness checks, not data-driven replacements.
The analysis uses audience tie draw 1 and Efron's tied-likelihood approximation,
matching the primary model.

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "arrow", "data.table", "digest", "jsonlite", "survival", "MASS", "knitr"
)
missing_packages <- required_packages[!vapply(
  required_packages, requireNamespace, logical(1), quietly = TRUE
)]
if (length(missing_packages)) {
  stop("Restore the R environment before knitting. Missing: ",
       paste(missing_packages, collapse = ", "))
}
library(data.table)
library(survival)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Configuration and sealed-test guard

```{r configuration}
feature_root <- Sys.getenv(
  "COMMENTGAP_MODEL_DATA_ROOT", "model_output/selection_2025/model_data"
)
regression_root <- Sys.getenv(
  "COMMENTGAP_REGRESSION_ROOT", "model_output/selection_2025/regression"
)

output_root <- Sys.getenv(
  "COMMENTGAP_COLLINEARITY_ROOT",
  "model_output/selection_2025/regression_sensitivity/collinearity_shared_preprocessing_v4"
)
cache_root <- file.path(output_root, "fit_cache")
split_path <- file.path(feature_root, "master_article_split.parquet")
parameters_path <- file.path(feature_root, "preprocessing_parameters.json")
manifest_path <- file.path(feature_root, "feature_manifest.json")
requested_scopes <- strsplit(
  Sys.getenv("COMMENTGAP_COLLINEARITY_SCOPES", "root,all"), ",", fixed = TRUE
)[[1L]]
requested_scopes <- trimws(requested_scopes)
run_cv <- tolower(Sys.getenv("COMMENTGAP_COLLINEARITY_RUN_CV", "1")) %in%
  c("1", "true", "yes")
bootstrap_draws <- as.integer(Sys.getenv(
  "COMMENTGAP_COLLINEARITY_BOOTSTRAP_DRAWS", "1000"
))
base_seed <- as.integer(Sys.getenv("COMMENTGAP_MODEL_SEED", "20260813"))
force_recompute <- tolower(Sys.getenv(
  "COMMENTGAP_COLLINEARITY_FORCE_RECOMPUTE", "0"
)) %in% c("1", "true", "yes")
high_correlation_threshold <- as.numeric(Sys.getenv(
  "COMMENTGAP_COLLINEARITY_CORRELATION_THRESHOLD", "0.70"
))

if (!all(requested_scopes %in% c("root", "all"))) {
  stop("COMMENTGAP_COLLINEARITY_SCOPES must contain only root and/or all")
}
if (!length(requested_scopes)) stop("At least one scope is required")
if (!is.finite(bootstrap_draws) || bootstrap_draws < 100L) {
  stop("COMMENTGAP_COLLINEARITY_BOOTSTRAP_DRAWS must be at least 100")
}
if (!file.exists(split_path) || !file.exists(manifest_path) || !file.exists(parameters_path)) {
  stop("Missing shared preprocessing artifacts; run 06A2_shared_model_preprocessing.ipynb first")
}
dir.create(output_root, recursive = TRUE, showWarnings = FALSE)
dir.create(cache_root, recursive = TRUE, showWarnings = FALSE)

article_split <- as.data.table(arrow::read_parquet(split_path))
required_split <- c("story_id", "split_role", "development_fold")
missing_split <- setdiff(required_split, names(article_split))
if (length(missing_split)) {
  stop("Shared split is missing: ", paste(missing_split, collapse = ", "))
}
article_split[, story_id := as.character(story_id)]
if (article_split[, anyDuplicated(story_id)]) stop("Duplicate article split keys")
if (!setequal(unique(article_split$split_role), c("development", "paper2_test"))) {
  stop("Expected development and paper2_test split roles")
}
development_split <- article_split[split_role == "development",
  .(story_id, development_fold)]
sealed_test_ids <- article_split[split_role == "paper2_test", story_id]
if (development_split[, any(development_fold < 0L)]) {
  stop("Development articles contain invalid fold assignments")
}
if (length(intersect(development_split$story_id, sealed_test_ids))) {
  stop("Development and sealed-test article IDs overlap")
}
expected_folds <- sort(unique(development_split$development_fold))
if (!identical(expected_folds, 0:4)) {
  stop("Expected the shared five development folds 0:4")
}

feature_manifest <- jsonlite::fromJSON(manifest_path, simplifyVector = FALSE)
preprocessing_parameters <- jsonlite::fromJSON(parameters_path, simplifyVector = FALSE)
if (!identical(as.integer(preprocessing_parameters$version), 4L)) {
  stop("Expected shared preprocessing version 4; force-rerun 06A2")
}
fold_feature_columns <- preprocessing_parameters$fold_feature_columns
if (!setequal(as.integer(names(fold_feature_columns)), expected_folds)) {
  stop("Fold-specific feature mappings do not match development folds")
}
aqua_features <- setdiff(
  grep("^aqua_.*_expected$", names(feature_manifest$features), value = TRUE),
  "aqua_score_expected"
)
if (length(aqua_features) != 20L) stop("Expected exactly 20 AQuA dimensions")

cat("Scopes:", paste(requested_scopes, collapse = ", "), "\n")
cat("Development articles:", nrow(development_split), "\n")
cat("Sealed test articles excluded:", length(sealed_test_ids), "\n")
cat("Five-fold development comparison:", run_cv, "\n")
```

# Helpers

```{r helpers}
elapsed_start <- Sys.time()
progress_message <- function(...) {
  elapsed <- as.numeric(difftime(Sys.time(), elapsed_start, units = "mins"))
  cat(sprintf("[%s; %.1f min] %s\n", format(Sys.time(), "%H:%M:%S"), elapsed,
              paste0(..., collapse = "")))
  flush.console()
}

file_signature <- function(path) {
  information <- file.info(path)
  list(
    path = normalizePath(path, mustWork = TRUE),
    size = unname(information$size),
    modified = format(information$mtime, tz = "UTC", usetz = TRUE)
  )
}

atomic_save_rds <- function(object, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  temporary <- tempfile(pattern = "fit-", tmpdir = dirname(path))
  on.exit(unlink(temporary), add = TRUE)
  saveRDS(object, temporary, compress = "gzip")
  if (!file.rename(temporary, path)) stop("Could not publish cache: ", path)
}

interaction_name <- function(coefficient_names, feature, interaction_variable) {
  candidates <- c(
    paste0(feature, ":", interaction_variable),
    paste0(interaction_variable, ":", feature)
  )
  found <- candidates[candidates %in% coefficient_names]
  if (length(found) != 1L) {
    stop("Cannot identify interaction coefficient for ", feature)
  }
  found[[1L]]
}

compact_fit <- function(fit, coding, features, variant) {
  model_covariance <- if (!is.null(fit$naive.var)) fit$naive.var else fit$var
  output <- list(
    coefficients = stats::coef(fit),
    robust_covariance = fit$var,
    model_covariance = model_covariance,
    loglik = fit$loglik,
    iterations = fit$iter,
    warnings = unique(attr(fit, "fit_warnings")),
    coding = coding,
    features = features,
    variant = variant
  )
  coefficient_names <- names(output$coefficients)
  dimnames(output$robust_covariance) <- list(coefficient_names, coefficient_names)
  dimnames(output$model_covariance) <- list(coefficient_names, coefficient_names)
  output
}

build_stacked <- function(choice_set, features, coding = "reference") {
  required <- c(
    "story_id", "comment_id", "n_picks", "curator_selected",
    "audience_selected_draw_01", features
  )
  missing <- setdiff(required, names(choice_set))
  if (length(missing)) stop("Choice set is missing: ", paste(missing, collapse = ", "))
  keep <- c("story_id", "comment_id", "n_picks", features)
  audience <- copy(choice_set[, ..keep])
  curator_rows <- copy(choice_set[, ..keep])
  audience[, selected := as.integer(choice_set$audience_selected_draw_01)]
  curator_rows[, selected := as.integer(choice_set$curator_selected)]
  if (coding == "reference") {
    audience[, curator := 0]
    curator_rows[, curator := 1]
  } else if (coding == "effect") {
    audience[, selector_code := -0.5]
    curator_rows[, selector_code := 0.5]
  } else {
    stop("Unknown selector coding: ", coding)
  }
  audience[, selector := "audience"]
  curator_rows[, selector := "curator"]
  stacked <- rbindlist(list(audience, curator_rows), use.names = TRUE, fill = TRUE)
  stacked[, stratum := paste(story_id, selector, sep = "::")]
  checks <- stacked[, .(
    selected_n = sum(selected), expected_n = first(n_picks), candidates_n = .N
  ), by = stratum]
  if (checks[, any(selected_n != expected_n)]) {
    stop("A development choice set does not contain exactly k selections")
  }
  if (checks[, any(selected_n <= 0L | selected_n >= candidates_n)]) {
    stop("A development choice set lacks outcome variation")
  }
  stacked
}

model_formula <- function(features, coding) {
  interaction_variable <- if (coding == "reference") "curator" else "selector_code"
  stats::as.formula(paste(
    "selected ~",
    paste(c(features, paste0(interaction_variable, ":", features),
            "strata(stratum)", "cluster(story_id)"), collapse = " + ")
  ))
}

fit_compact <- function(
    choice_set, features, coding, scope, variant, fold = "full",
    allow_convergence_warnings = FALSE
) {
  cache_path <- file.path(
    cache_root, scope, variant, paste0("fold_", fold, ".rds")
  )
  fingerprint <- digest::digest(list(
    schema = "collinearity-sensitivity-shared-preprocessing-v4",
    scope = scope,
    variant = variant,
    fold = as.character(fold),
    coding = coding,
    features = features,
    stories = sort(unique(choice_set$story_id)),
    split = file_signature(split_path),
    input = file_signature(file.path(
      feature_root, paste0("choice_set_", scope, ".parquet"))),
    manifest = file_signature(manifest_path),
    survival = as.character(utils::packageVersion("survival")),
    r = paste(R.version$major, R.version$minor, sep = ".")
  ), algo = "sha256")
  if (!force_recompute && file.exists(cache_path)) {
    cached <- readRDS(cache_path)
    if (identical(cached$fingerprint, fingerprint)) {
      cached_warnings <- if (is.null(cached$fit$warnings)) {
        character()
      } else {
        unique(cached$fit$warnings)
      }
      cached_serious <- grep(
        "infinite|did not converge|ran out of iterations",
        cached_warnings, ignore.case = TRUE, value = TRUE
      )
      if (!length(cached_serious) || allow_convergence_warnings) {
        progress_message("CACHE HIT ", scope, "/", variant, " fold=", fold)
        return(cached$fit)
      }
      progress_message("CACHE WARNING REJECTED ", scope, "/", variant,
                       " fold=", fold, " under strict convergence mode")
    }
  }
  progress_message("FITTING ", scope, "/", variant, " fold=", fold,
                   " rows=", format(2L * nrow(choice_set), big.mark = ","))
  stacked <- build_stacked(choice_set, features, coding)
  fit_warnings <- character()
  fit <- withCallingHandlers(
    survival::clogit(
      model_formula(features, coding), data = stacked, method = "efron",
      model = FALSE, x = FALSE, y = FALSE
    ),
    warning = function(condition) {
      fit_warnings <<- c(fit_warnings, conditionMessage(condition))
      invokeRestart("muffleWarning")
    }
  )
  attr(fit, "fit_warnings") <- unique(fit_warnings)
  if (anyNA(stats::coef(fit)) || any(!is.finite(stats::coef(fit)))) {
    stop("Unidentified or non-finite coefficients in ", scope, "/", variant)
  }
  if (any(!is.finite(fit$var)) ||
      (!is.null(fit$naive.var) && any(!is.finite(fit$naive.var)))) {
    stop("Non-finite covariance matrix in ", scope, "/", variant)
  }
  serious <- grep(
    "infinite|did not converge|ran out of iterations", fit_warnings,
    ignore.case = TRUE, value = TRUE
  )
  if (length(serious) && !allow_convergence_warnings) {
    stop("Convergence failure: ", paste(serious, collapse = " | "))
  }
  if (length(serious)) {
    progress_message(
      "WARNING RETAINED ", scope, "/", variant, " fold=", fold, ": ",
      paste(serious, collapse = " | ")
    )
  }
  result <- compact_fit(fit, coding, features, variant)
  atomic_save_rds(list(fingerprint = fingerprint, fit = result), cache_path)
  rm(stacked, fit)
  gc(verbose = FALSE)
  result
}

prepare_features <- function(training, validation, features, standardize) {
  training <- copy(training)
  validation <- if (is.null(validation)) NULL else copy(validation)
  working_features <- as.character(features)
  if (is.null(names(standardize)) || any(!nzchar(names(standardize)))) {
    stop("Standardization contract must be a named logical vector")
  }
  missing_standardize <- setdiff(working_features, names(standardize))
  if (length(missing_standardize)) {
    stop("Missing standardization flags for active feature(s): ",
         paste(missing_standardize, collapse = ", "))
  }
  working_standardize <- standardize[working_features]
  scaling <- rbindlist(lapply(working_features, function(feature) {
    do_scale <- isTRUE(unname(working_standardize[feature]))
    if (!do_scale) {
      return(data.table(term = feature, mean = NA_real_, sd = NA_real_))
    }
    center <- mean(as.numeric(training[[feature]]))
    spread <- stats::sd(as.numeric(training[[feature]]))
    if (!is.finite(spread) || spread <= 0) stop("No variation in ", feature)
    set(training, j = feature,
        value = (as.numeric(training[[feature]]) - center) / spread)
    if (!is.null(validation)) {
      set(validation, j = feature,
          value = (as.numeric(validation[[feature]]) - center) / spread)
    }
    data.table(term = feature, mean = center, sd = spread)
  }))
  if (training[, anyNA(.SD), .SDcols = working_features] ||
      (!is.null(validation) && validation[, anyNA(.SD), .SDcols = working_features])) {
    stop("Missing values remain after development-only scaling")
  }
  list(
    training = training,
    validation = validation,
    features = working_features,
    scaling = scaling
  )
}

selector_coefficients <- function(fit) {
  coefficient_names <- names(fit$coefficients)
  interaction_variable <- if (fit$coding == "reference") "curator" else "selector_code"
  rbindlist(lapply(fit$features, function(feature) {
    interaction <- interaction_name(coefficient_names, feature, interaction_variable)
    main <- unname(fit$coefficients[[feature]])
    difference <- unname(fit$coefficients[[interaction]])
    if (fit$coding == "reference") {
      audience <- main
      curator_value <- main + difference
    } else {
      audience <- main - 0.5 * difference
      curator_value <- main + 0.5 * difference
    }
    data.table(
      term = feature, audience = audience, curator = curator_value,
      curator_minus_audience = difference
    )
  }))
}

score_compact <- function(fit, choice_set) {
  associations <- selector_coefficients(fit)
  features <- fit$features
  matrix_x <- as.matrix(choice_set[, ..features])
  audience_score <- drop(matrix_x %*% associations$audience)
  curator_score <- drop(matrix_x %*% associations$curator)
  rbindlist(list(
    choice_set[, .(
      story_id, comment_id, development_fold, n_picks,
      selector = "audience", selected = as.integer(audience_selected_draw_01),
      score = audience_score
    )],
    choice_set[, .(
      story_id, comment_id, development_fold, n_picks,
      selector = "curator", selected = as.integer(curator_selected),
      score = curator_score
    )]
  ))
}

evaluate_scores <- function(scores) {
  scores[, {
    k <- first(n_picks)
    ranked <- order(-score, comment_id)
    ideal <- order(-selected, comment_id)
    predicted <- ranked[seq_len(k)]
    actual <- which(selected == 1L)
    discounts <- 1 / log2(seq_len(k) + 1)
    dcg <- sum(selected[predicted] * discounts)
    idcg <- sum(selected[ideal[seq_len(k)]] * discounts)
    overlap <- length(intersect(predicted, actual))
    .(
      development_fold = first(development_fold),
      n_candidates = .N, n_picks = k,
      ndcg_at_k = dcg / idcg,
      top_k_overlap = overlap / k,
      jaccard = overlap / length(union(predicted, actual))
    )
  }, by = .(story_id, selector)]
}

make_variants <- function(scope, full_features) {
  remove <- function(x) setdiff(full_features, x)
  variants <- list(
    full_reference = list(features = full_features, coding = "reference"),
    no_author_history = list(
      features = remove(c(
        "log_author_prior_30d_comments",
        "author_prior_30d_upvote_reception",
        "author_prior_30d_downvote_reception",
        "log_author_prior_comments_story"
      )), coding = "reference"
    ),
    no_aqua = list(features = remove(aqua_features), coding = "reference")
  )
  if (scope == "all") {
    variants <- c(variants, list(
      reply_indicator_only = list(
        features = remove("reply_depth_centered"), coding = "reference"
      ),
      reply_depth_only = list(
        features = remove("is_reply"), coding = "reference"
      ),
      reduced_activity = list(
        features = remove(c(
          "discussion_pace", "prior_reply_composition"
        )), coding = "reference"
      )
    ))
  }
  variants
}

feature_blocks <- function(scope, features) {
  blocks <- list(
    global_activity = intersect(features, c(
      "log_hours_since_article", "prior_reply_composition", "discussion_pace"
    )),
    author_history = intersect(features, c(
      "log_author_prior_30d_comments",
      "author_prior_30d_upvote_reception",
      "author_prior_30d_downvote_reception",
      "log_author_prior_comments_story"
    )),
    aqua_dimensions = intersect(features, aqua_features),
    sentiment_and_safety = intersect(features, c(
      "sentiment_positive", "sentiment_negative", "toxicity_probability"
    ))
  )
  if (scope == "all") {
    blocks <- c(list(
      reply_structure = intersect(features, c("is_reply", "reply_depth_centered"))
    ), blocks)
  }
  blocks[vapply(blocks, length, integer(1)) > 0L]
}

wald_block_test <- function(fit, block, family) {
  coefficient_names <- names(fit$coefficients)
  p <- length(coefficient_names)
  interaction_variable <- if (fit$coding == "reference") "curator" else "selector_code"
  rows <- list()
  if (family == "all_parameters") {
    for (feature in block) {
      interaction <- interaction_name(coefficient_names, feature, interaction_variable)
      first <- numeric(p); names(first) <- coefficient_names; first[[feature]] <- 1
      second <- numeric(p); names(second) <- coefficient_names; second[[interaction]] <- 1
      rows <- c(rows, list(first, second))
    }
  } else {
    for (feature in block) {
      interaction <- interaction_name(coefficient_names, feature, interaction_variable)
      row <- numeric(p); names(row) <- coefficient_names
      if (family == "audience") row[[feature]] <- 1
      if (family == "difference") row[[interaction]] <- 1
      if (family == "curator") {
        row[[feature]] <- 1
        row[[interaction]] <- 1
      }
      rows <- c(rows, list(row))
    }
  }
  contrast <- do.call(rbind, rows)
  estimate <- drop(contrast %*% fit$coefficients)
  covariance <- contrast %*% fit$robust_covariance %*% t(contrast)
  degrees <- qr(covariance)$rank
  statistic <- drop(t(estimate) %*% MASS::ginv(covariance) %*% estimate)
  data.table(
    family = family, terms = length(block), df = degrees,
    chisq = statistic,
    p_value = stats::pchisq(statistic, df = degrees, lower.tail = FALSE)
  )
}

within_story_gram <- function(choice_set, features, coding) {
  x <- as.matrix(choice_set[, ..features])
  storage.mode(x) <- "double"
  if (any(!is.finite(x))) stop("Non-finite design values")
  story_factor <- match(choice_set$story_id, unique(choice_set$story_id))
  group_sums <- rowsum(x, story_factor, reorder = FALSE)
  group_n <- tabulate(story_factor)
  centered_gram <- crossprod(x) - crossprod(group_sums / sqrt(group_n))
  centered_gram <- (centered_gram + t(centered_gram)) / 2
  p <- length(features)
  if (coding == "reference") {
    expanded <- rbind(cbind(2 * centered_gram, centered_gram),
                      cbind(centered_gram, centered_gram))
    interaction_suffix <- ":curator"
  } else {
    zero <- matrix(0, p, p)
    expanded <- rbind(cbind(2 * centered_gram, zero),
                      cbind(zero, 0.5 * centered_gram))
    interaction_suffix <- ":selector_code"
  }
  terms <- c(features, paste0(features, interaction_suffix))
  dimnames(expanded) <- list(terms, terms)
  expanded
}

gram_diagnostics <- function(gram, scope, variant) {
  diagonal <- diag(gram)
  if (any(!is.finite(diagonal) | diagonal <= .Machine$double.eps)) {
    bad <- names(diagonal)[!is.finite(diagonal) | diagonal <= .Machine$double.eps]
    stop("No within-story information for: ", paste(bad, collapse = ", "))
  }
  correlation <- gram / sqrt(outer(diagonal, diagonal))
  correlation <- (correlation + t(correlation)) / 2
  eig <- eigen(correlation, symmetric = TRUE)
  tolerance <- max(eig$values) * .Machine$double.eps * nrow(correlation)
  safe_values <- pmax(eig$values, tolerance)
  condition_index <- sqrt(max(safe_values) / safe_values)
  components <- sweep(eig$vectors^2, 2L, safe_values, "/")
  proportions <- components / rowSums(components)
  dimension_names <- paste0("dimension_", seq_along(safe_values))
  dimnames(proportions) <- list(rownames(correlation), dimension_names)
  condition <- data.table(
    scope = scope, variant = variant,
    dimension = dimension_names,
    eigenvalue = eig$values,
    condition_index = condition_index
  )
  variance <- as.data.table(as.table(proportions))
  setnames(variance, c("term", "dimension", "variance_proportion"))
  variance[, `:=`(scope = scope, variant = variant)]
  upper <- which(upper.tri(correlation), arr.ind = TRUE)
  pairs <- data.table(
    scope = scope, variant = variant,
    term_1 = rownames(correlation)[upper[, 1L]],
    term_2 = colnames(correlation)[upper[, 2L]],
    correlation = correlation[upper]
  )
  pairs <- pairs[order(-abs(correlation))]
  list(
    correlation = correlation,
    condition = condition,
    variance = variance,
    pairs = pairs[abs(correlation) >= high_correlation_threshold]
  )
}

covariance_diagnostics <- function(fit, scope) {
  rbindlist(lapply(c("model_based", "cluster_robust"), function(kind) {
    covariance <- if (kind == "model_based") {
      fit$model_covariance
    } else {
      fit$robust_covariance
    }
    covariance <- (covariance + t(covariance)) / 2
    covariance_correlation <- stats::cov2cor(covariance)
    values <- eigen(
      covariance_correlation, symmetric = TRUE, only.values = TRUE
    )$values
    positive <- values[values > max(values) * .Machine$double.eps * length(values)]
    data.table(
      scope = scope, covariance = kind,
      raw_covariance_condition_number = kappa(covariance, exact = TRUE),
      correlation_scale_condition_number = kappa(
        covariance_correlation, exact = TRUE
      ),
      corresponding_max_condition_index = sqrt(max(positive) / min(positive)),
      coefficients = length(fit$coefficients)
    )
  }))
}

coefficient_correlation_pairs <- function(fit, scope) {
  rbindlist(lapply(c("model_based", "cluster_robust"), function(kind) {
    covariance <- if (kind == "model_based") {
      fit$model_covariance
    } else {
      fit$robust_covariance
    }
    correlation <- stats::cov2cor(covariance)
    upper <- which(upper.tri(correlation), arr.ind = TRUE)
    output <- data.table(
      scope = scope, covariance = kind,
      term_1 = rownames(correlation)[upper[, 1L]],
      term_2 = colnames(correlation)[upper[, 2L]],
      correlation = correlation[upper]
    )
    output[abs(correlation) >= high_correlation_threshold][
      order(-abs(correlation))]
  }))
}

paired_cv_summary <- function(metrics, full_name, draws, seed) {
  metric_names <- c("ndcg_at_k", "top_k_overlap", "jaccard")
  full <- metrics[variant == full_name]
  comparisons <- setdiff(unique(metrics$variant), full_name)
  rbindlist(lapply(seq_along(comparisons), function(variant_index) {
    variant_name <- comparisons[[variant_index]]
    candidate <- metrics[variant == variant_name]
    merged <- merge(
      candidate, full,
      by = c("scope", "story_id", "selector", "development_fold"),
      suffixes = c("_variant", "_full"), sort = FALSE
    )
    if (nrow(merged) != nrow(full)) {
      stop("CV variants do not contain identical article-selector keys")
    }
    rbindlist(lapply(seq_along(metric_names), function(metric_index) {
      metric <- metric_names[[metric_index]]
      rbindlist(lapply(c("audience", "curator"), function(selector_name) {
        subset <- merged[selector == selector_name]
        difference <- subset[[paste0(metric, "_variant")]] -
          subset[[paste0(metric, "_full")]]
        set.seed(seed + variant_index * 10000L + metric_index * 100L +
                   match(selector_name, c("audience", "curator")))
        bootstrap <- replicate(
          draws, mean(sample(difference, length(difference), replace = TRUE))
        )
        interval <- unname(stats::quantile(bootstrap, c(0.025, 0.975)))
        data.table(
          variant = variant_name, selector = selector_name, metric = metric,
          full_estimate = mean(subset[[paste0(metric, "_full")]]),
          variant_estimate = mean(subset[[paste0(metric, "_variant")]]),
          variant_minus_full = mean(difference),
          conf_low = interval[[1L]], conf_high = interval[[2L]],
          articles = nrow(subset), bootstrap_draws = draws
        )
      }))
    }))
  }))
}
```

# Run one scope

```{r run-scope}
run_scope <- function(scope) {
  progress_message("Starting scope=", scope)
  input_path <- file.path(feature_root, paste0("choice_set_", scope, ".parquet"))
  primary_path <- file.path(regression_root, scope, "stacked_selection_model.rds")
  if (!file.exists(input_path) || !file.exists(primary_path)) {
    stop("Missing choice set or primary 06B fit for ", scope)
  }
  full_features <- unlist(
    feature_manifest$models[[scope]]$features, use.names = FALSE
  )
  missing_metadata <- setdiff(full_features, names(feature_manifest$features))
  if (length(missing_metadata)) {
    stop("Missing feature metadata for active feature(s): ",
         paste(missing_metadata, collapse = ", "))
  }
  standardize <- setNames(vapply(
    full_features,
    function(feature) isTRUE(feature_manifest$features[[feature]]$standardize),
    logical(1)
  ), full_features)
  active_fold_mappings <- lapply(fold_feature_columns, function(mapping) {
    mapping[intersect(names(mapping), full_features)]
  })
  fold_source_columns <- if (run_cv) unique(unlist(active_fold_mappings, use.names = FALSE)) else character()
  required_columns <- c(
    "story_id", "comment_id", "n_picks", "curator_selected",
    "audience_selected_draw_01", full_features, fold_source_columns
  )
  source <- as.data.table(arrow::read_parquet(input_path, col_select = required_columns))
  source[, story_id := as.character(story_id)]
  if (source[, anyDuplicated(paste(story_id, comment_id, sep = "::"))]) {
    stop("Duplicate candidate keys in ", scope)
  }
  # The source parquet contains both partitions. Only development rows survive
  # this merge, and every downstream object is checked against the sealed IDs.
  development <- merge(
    source, development_split, by = "story_id", all = FALSE, sort = FALSE
  )
  rm(source)
  gc(verbose = FALSE)
  if (uniqueN(development$story_id) != nrow(development_split)) {
    stop(scope, " does not contain every development article")
  }
  if (development[, any(story_id %in% sealed_test_ids)]) {
    stop("SEALED-TEST GUARD: test rows entered the development object")
  }
  if (!setequal(unique(development$development_fold), expected_folds)) {
    stop("Development fold mismatch in ", scope)
  }

  prepared_full <- prepare_features(
    development, NULL, full_features, standardize
  )
  primary_large <- readRDS(primary_path)
  primary <- compact_fit(primary_large, "reference", full_features, "full_reference")
  rm(primary_large)
  gc(verbose = FALSE)
  expected_interactions <- vapply(
    full_features,
    function(feature) interaction_name(
      names(primary$coefficients), feature, "curator"
    ), character(1)
  )
  if (!all(c(full_features, expected_interactions) %in%
           names(primary$coefficients))) {
    stop("Saved primary model does not match the feature contract")
  }

  covariance_summary <- covariance_diagnostics(primary, scope)
  coefficient_pairs <- coefficient_correlation_pairs(primary, scope)
  blocks <- feature_blocks(scope, full_features)
  block_tests <- rbindlist(lapply(names(blocks), function(block_name) {
    output <- rbindlist(lapply(
      c("audience", "curator", "difference", "all_parameters"),
      function(family) wald_block_test(primary, blocks[[block_name]], family)
    ))
    output[, `:=`(scope = scope, block = block_name)]
    output
  }))
  block_tests[, q_value_bh := stats::p.adjust(p_value, method = "BH"),
              by = family]

  ordinary_variants <- make_variants(scope, full_features)
  analysis_variants <- ordinary_variants
  analysis_variants$selector_effect_coding_exact <- list(
    features = full_features, coding = "effect"
  )
  invalid_variants <- names(Filter(
    function(specification) {
      length(setdiff(specification$features, full_features)) > 0L ||
        anyDuplicated(specification$features) > 0L
    },
    analysis_variants
  ))
  if (length(invalid_variants)) {
    stop("Sensitivity variants must use unique subsets of primary features: ",
         paste(invalid_variants, collapse = ", "))
  }

  association_rows <- list()
  fit_rows <- list()
  condition_rows <- list()
  variance_rows <- list()
  design_pair_rows <- list()
  exact_rows <- list()
  full_scores <- score_compact(primary, prepared_full$training)

  for (variant_name in names(analysis_variants)) {
    specification <- analysis_variants[[variant_name]]
    if (variant_name == "full_reference") {
      prepared <- prepared_full
      fit <- primary
    } else {
      prepared <- prepare_features(
        development, NULL, specification$features, standardize
      )
      fit <- fit_compact(
        prepared$training, prepared$features, specification$coding,
        scope, variant_name
      )
    }
    association <- selector_coefficients(fit)
    association[, `:=`(
      scope = scope, variant = variant_name,
      comparison_term = term
    )]
    association_rows[[variant_name]] <- association
    fit_rows[[variant_name]] <- data.table(
      scope = scope, variant = variant_name, coding = fit$coding,
      parameters = length(fit$coefficients),
      fitted_log_likelihood = fit$loglik[[2L]],
      iterations = paste(fit$iterations, collapse = ","),
      warning_count = length(fit$warnings)
    )

    design <- gram_diagnostics(
      within_story_gram(prepared$training, prepared$features, specification$coding),
      scope, variant_name
    )
    condition_rows[[variant_name]] <- design$condition
    variance_rows[[variant_name]] <- design$variance
    design_pair_rows[[variant_name]] <- design$pairs

    if (grepl("_exact$", variant_name)) {
      variant_scores <- score_compact(fit, prepared$training)
      comparison <- merge(
        variant_scores[, .(story_id, comment_id, selector, variant_score = score)],
        full_scores[, .(story_id, comment_id, selector, full_score = score)],
        by = c("story_id", "comment_id", "selector"), sort = FALSE
      )
      comparison[, score_difference := variant_score - full_score]
      comparison[, centred_difference :=
        score_difference - mean(score_difference), by = .(story_id, selector)]
      rank_check <- comparison[, .(
        same = identical(
          order(-full_score, comment_id), order(-variant_score, comment_id)
        )
      ), by = .(story_id, selector)]
      exact_rows[[variant_name]] <- data.table(
        scope = scope, variant = variant_name,
        log_likelihood_difference = fit$loglik[[2L]] - primary$loglik[[2L]],
        maximum_absolute_within_stratum_score_difference =
          max(abs(comparison$centred_difference)),
        ranking_identical = all(rank_check$same)
      )
      rm(variant_scores, comparison, rank_check)
    }
    rm(prepared, fit, design)
    gc(verbose = FALSE)
  }

  associations <- rbindlist(association_rows, use.names = TRUE, fill = TRUE)
  fit_summaries <- rbindlist(fit_rows, use.names = TRUE, fill = TRUE)
  design_conditions <- rbindlist(condition_rows)
  design_variance <- rbindlist(variance_rows)
  design_pairs <- rbindlist(design_pair_rows, use.names = TRUE, fill = TRUE)
  exact_checks <- rbindlist(exact_rows, use.names = TRUE, fill = TRUE)

  full_associations <- associations[variant == "full_reference",
    .(comparison_term, full_audience = audience, full_curator = curator,
      full_difference = curator_minus_audience)]
  coefficient_stability <- merge(
    associations[!grepl("_exact$", variant),
      .(scope, variant, comparison_term, audience, curator,
        difference = curator_minus_audience)],
    full_associations, by = "comparison_term", all.x = TRUE, sort = FALSE
  )
  coefficient_stability <- melt(
    coefficient_stability,
    id.vars = c("scope", "variant", "comparison_term"),
    measure.vars = list(
      estimate = c("audience", "curator", "difference"),
      full_estimate = c("full_audience", "full_curator", "full_difference")
    ),
    variable.name = "estimand"
  )
  coefficient_stability[, estimand := c(
    "audience", "curator", "curator_minus_audience"
  )[estimand]]
  coefficient_stability[, `:=`(
    estimate_change = estimate - full_estimate,
    absolute_estimate_change = abs(estimate - full_estimate),
    relative_absolute_change = fifelse(
      abs(full_estimate) >= 0.05,
      abs(estimate - full_estimate) / abs(full_estimate), NA_real_
    ),
    sign_change = sign(estimate) != sign(full_estimate),
    material_sign_change = sign(estimate) != sign(full_estimate) &
      abs(full_estimate) >= 0.05 & abs(estimate) >= 0.05
  )]

  cv_metrics <- data.table()
  cv_summary <- data.table()
  cv_differences <- data.table()
  cv_fit_warnings <- data.table(
    scope = character(), variant = character(), fold = integer(),
    warning_count = integer(), warnings = character(),
    maximum_absolute_coefficient = numeric()
  )
  if (run_cv) {
    cv_rows <- list()
    cv_warning_rows <- list()
    for (variant_name in names(ordinary_variants)) {
      specification <- ordinary_variants[[variant_name]]
      for (fold in expected_folds) {
        training_raw <- development[development_fold != fold]
        validation_raw <- development[development_fold == fold]
        replacements <- active_fold_mappings[[as.character(fold)]]
        replacements <- replacements[
          intersect(names(replacements), specification$features)
        ]
        for (feature in names(replacements)) {
          fold_column <- unname(unlist(replacements[[feature]]))
          if (!fold_column %in% names(development)) stop("Missing ", fold_column)
          set(training_raw, j = feature, value = training_raw[[fold_column]])
          set(validation_raw, j = feature, value = validation_raw[[fold_column]])
        }
        if (training_raw[, any(story_id %in% validation_raw$story_id)]) {
          stop("CV training-validation article overlap")
        }
        prepared <- prepare_features(
          training_raw, validation_raw, specification$features, standardize
        )
        fit <- fit_compact(
          prepared$training, prepared$features, specification$coding,
          scope, variant_name, fold = fold,
          allow_convergence_warnings = TRUE
        )
        scores <- score_compact(fit, prepared$validation)
        if (any(!is.finite(scores$score))) {
          stop("Non-finite validation scores in ", scope, "/", variant_name,
               " fold=", fold)
        }
        fold_metrics <- evaluate_scores(scores)
        warning_text <- paste(fit$warnings, collapse = " | ")
        fold_metrics[, `:=`(
          scope = scope, variant = variant_name,
          fit_warning_count = length(fit$warnings),
          fit_warning = warning_text
        )]
        if (length(fit$warnings)) {
          cv_warning_rows[[paste(variant_name, fold, sep = "::")]] <- data.table(
            scope = scope, variant = variant_name, fold = as.integer(fold),
            warning_count = length(fit$warnings), warnings = warning_text,
            maximum_absolute_coefficient = max(abs(fit$coefficients))
          )
        }
        cv_rows[[paste(variant_name, fold, sep = "::")]] <- fold_metrics
        rm(training_raw, validation_raw, prepared, fit, scores, fold_metrics)
        gc(verbose = FALSE)
      }
    }
    if (length(cv_warning_rows)) cv_fit_warnings <- rbindlist(cv_warning_rows)
    cv_metrics <- rbindlist(cv_rows)
    metric_long <- melt(
      cv_metrics,
      id.vars = c("scope", "variant", "story_id", "selector", "development_fold"),
      measure.vars = c("ndcg_at_k", "top_k_overlap", "jaccard"),
      variable.name = "metric", value.name = "value"
    )
    cv_summary <- metric_long[, {
      set.seed(base_seed + .GRP)
      estimates <- replicate(
        bootstrap_draws, mean(sample(value, .N, replace = TRUE))
      )
      interval <- unname(stats::quantile(estimates, c(0.025, 0.975)))
      .(estimate = mean(value), conf_low = interval[[1L]],
        conf_high = interval[[2L]], articles = .N,
        bootstrap_draws = bootstrap_draws)
    }, by = .(scope, variant, selector, metric)]
    cv_differences <- paired_cv_summary(
      cv_metrics, "full_reference", bootstrap_draws,
      base_seed + if (scope == "root") 100000L else 200000L
    )
    cv_differences[, scope := scope]
  }

  condition_summary <- design_conditions[, .(
    maximum_condition_index = max(condition_index),
    dimensions_over_30 = sum(condition_index >= 30),
    dimensions_over_15 = sum(condition_index >= 15)
  ), by = .(scope, variant)]

  scope_output <- file.path(output_root, scope)
  dir.create(scope_output, recursive = TRUE, showWarnings = FALSE)
  fwrite(covariance_summary, file.path(scope_output, "covariance_conditioning.csv"))
  fwrite(coefficient_pairs,
         file.path(scope_output, "high_coefficient_correlations.csv"))
  fwrite(block_tests, file.path(scope_output, "joint_block_wald_tests.csv"))
  fwrite(fit_summaries, file.path(scope_output, "variant_fit_summary.csv"))
  fwrite(associations, file.path(scope_output, "variant_selector_coefficients.csv"))
  fwrite(coefficient_stability,
         file.path(scope_output, "variant_coefficient_stability.csv"))
  fwrite(design_conditions, file.path(scope_output, "design_condition_indices.csv"))
  fwrite(condition_summary, file.path(scope_output, "design_condition_summary.csv"))
  fwrite(design_pairs, file.path(scope_output, "high_design_correlations.csv"))
  arrow::write_parquet(
    design_variance, file.path(scope_output, "design_variance_decomposition.parquet")
  )
  fwrite(exact_checks,
         file.path(scope_output, "exact_reparameterisation_checks.csv"))
  if (run_cv) {
    arrow::write_parquet(
      cv_metrics, file.path(scope_output, "development_cv_article_metrics.parquet")
    )
    fwrite(cv_fit_warnings,
           file.path(scope_output, "development_cv_fit_warnings.csv"))
    fwrite(cv_summary, file.path(scope_output, "development_cv_metric_summary.csv"))
    fwrite(cv_differences,
           file.path(scope_output, "development_cv_paired_differences.csv"))
  }
  jsonlite::write_json(list(
    analysis = "development-only collinearity sensitivity",
    scope = scope,
    shared_preprocessing_version = as.integer(preprocessing_parameters$version),
    feature_manifest = file_signature(manifest_path),
    created_at = format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z"),
    development_articles = uniqueN(development$story_id),
    sealed_test_articles_excluded = length(sealed_test_ids),
    sealed_test_performance_read = FALSE,
    audience_tie_draw = 1L, method = "efron",
    cv_enabled = run_cv, cv_folds = expected_folds,
    bootstrap_draws = if (run_cv) bootstrap_draws else 0L,
    cv_warning_fits = nrow(cv_fit_warnings),
    variants = names(analysis_variants),
    primary_model_status = "fixed reference; reduced variants are sensitivities"
  ), file.path(scope_output, "analysis_manifest.json"),
  auto_unbox = TRUE, pretty = TRUE)

  progress_message("Completed scope=", scope)
  list(
    scope = scope, covariance = covariance_summary,
    coefficient_pairs = coefficient_pairs, block_tests = block_tests,
    condition_summary = condition_summary, exact_checks = exact_checks,
    cv_summary = cv_summary, cv_differences = cv_differences,
    cv_fit_warnings = cv_fit_warnings
  )
}

results <- lapply(requested_scopes, run_scope)
progress_message("Collinearity sensitivity completed successfully")
```

# Results

```{r results, echo=FALSE}
for (result in results) {
  cat("\n##", if (result$scope == "root") "Root candidates" else "All comments",
      "\n\n")
  print(knitr::kable(
    result$covariance, digits = 2,
    caption = "Raw and correlation-scale covariance conditioning"
  ))
  cat("\n### Strongest coefficient correlations\n\n")
  print(knitr::kable(
    head(result$coefficient_pairs, 20L), digits = 3,
    caption = "Coefficient correlations above the configured threshold"
  ))
  cat("\n### Within-choice-set design conditioning\n\n")
  print(knitr::kable(
    result$condition_summary, digits = 2,
    caption = "Condition indices for the expanded within-story-centred design"
  ))
  if (nrow(result$exact_checks)) {
    cat("\n### Equivalent parameterisation checks\n\n")
    print(knitr::kable(
      result$exact_checks, digits = 8,
      caption = "Likelihood and ranking equivalence checks"
    ))
  }
  cat("\n### Joint block tests\n\n")
  print(knitr::kable(
    result$block_tests, digits = 4,
    caption = "Cluster-robust development-set joint Wald tests"
  ))
  if (run_cv) {
    if (nrow(result$cv_fit_warnings)) {
      cat("\n### CV fit warnings retained for review\n\n")
      print(knitr::kable(result$cv_fit_warnings, digits = 4))
    }
    cat("\n### Five-fold development performance\n\n")
    print(knitr::kable(result$cv_summary, digits = 4))
    cat("\n### Paired reduced-minus-full comparisons\n\n")
    print(knitr::kable(
      result$cv_differences, digits = 4,
      caption = "Positive values favour the sensitivity model"
    ))
  }
}
```

# Interpretation rules

The unscaled covariance condition number is retained for continuity with 06B,
but conclusions should rely on correlation-scale conditioning, within-choice-set
condition indices, variance-decomposition proportions, and coefficient stability
together.

A condition index above 30 prompts inspection; it is not an automatic model
failure. Stronger evidence is a high-index dimension on which several terms have
large variance-decomposition proportions. Such terms should be interpreted
jointly, especially if individual estimates change across reduced models.

Selector effect coding and centred reply depth are exact reparameterisations.
If their likelihoods and within-stratum rankings agree, they may replace the
presentation parameterisation without constituting a new model choice. Other
variants omit information. They remain post-hoc development sensitivities, and
the already-opened Paper 2 test partition must not be used to choose among them.
