---
title: "Stacked curator–audience selection models (2025)"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Objective

Fit two common-scale conditional-logit models: one whose candidates are root
comments and one whose candidates are all published comments. Each comment is
stacked once for the audience selector and once for the curator selector.
Article-by-selector strata provide the article controls; standard errors are
clustered by article.

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "arrow", "data.table", "jsonlite", "survival", "MASS", "knitr"
)
missing_packages <- required_packages[!vapply(
  required_packages, requireNamespace, logical(1), quietly = TRUE
)]
if (length(missing_packages)) {
  stop(
    "Restore the R environment before knitting. Missing: ",
    paste(missing_packages, collapse = ", ")
  )
}
library(data.table)
library(survival)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Configuration and shared feature contract

```{r configuration}
feature_root <- Sys.getenv(
  "COMMENTGAP_FEATURE_ROOT",
  "model_output/selection_2025/features"
)
output_root <- Sys.getenv(
  "COMMENTGAP_REGRESSION_ROOT",
  "model_output/selection_2025/regression"
)
tie_draws_requested <- as.integer(Sys.getenv("COMMENTGAP_TIE_DRAWS", "10"))
probability_draws <- as.integer(Sys.getenv("COMMENTGAP_PROBABILITY_DRAWS", "5000"))
base_seed <- as.integer(Sys.getenv("COMMENTGAP_MODEL_SEED", "20260813"))

if (tie_draws_requested < 1L) stop("COMMENTGAP_TIE_DRAWS must be positive")
if (probability_draws < 100L) stop("COMMENTGAP_PROBABILITY_DRAWS must be at least 100")
dir.create(output_root, recursive = TRUE, showWarnings = FALSE)

manifest_path <- file.path(feature_root, "feature_manifest.json")
provenance_path <- file.path(feature_root, "provenance_manifest.json")
if (!file.exists(manifest_path)) stop("Missing feature manifest: ", manifest_path)
if (!file.exists(provenance_path)) stop("Missing provenance manifest: ", provenance_path)
feature_manifest <- jsonlite::fromJSON(manifest_path, simplifyVector = FALSE)
provenance <- jsonlite::fromJSON(provenance_path, simplifyVector = FALSE)
cat("Feature watermark:", provenance$watermark, "\n")
```

# Model helpers

```{r helpers}
safe_scale <- function(x, name) {
  x <- as.numeric(x)
  center <- mean(x)
  spread <- stats::sd(x)
  if (!is.finite(spread) || spread <= 0) {
    stop("Cannot standardize feature with no variation: ", name)
  }
  list(value = (x - center) / spread, mean = center, sd = spread)
}

interaction_name <- function(fit, feature) {
  candidates <- c(paste0(feature, ":curator"), paste0("curator:", feature))
  found <- candidates[candidates %in% names(stats::coef(fit))]
  if (length(found) != 1L) stop("Cannot identify interaction for ", feature)
  found[[1L]]
}

build_stacked <- function(choice_set, features, draw) {
  label <- sprintf("audience_selected_draw_%02d", draw)
  if (!label %in% names(choice_set)) stop("Missing tie-draw label: ", label)
  keep <- c("story_id", "comment_id", "n_candidates", "n_picks", features)

  audience <- copy(choice_set[, ..keep])
  audience[, `:=`(
    selector = "audience",
    curator = 0L,
    selected = as.integer(choice_set[[label]])
  )]
  curator_rows <- copy(choice_set[, ..keep])
  curator_rows[, `:=`(
    selector = "curator",
    curator = 1L,
    selected = as.integer(choice_set$curator_selected)
  )]
  stacked <- rbindlist(list(audience, curator_rows), use.names = TRUE)
  stacked[, stratum := paste(story_id, selector, sep = "::")]
  setorder(stacked, stratum, comment_id)

  checks <- stacked[, .(
    selected_n = sum(selected),
    expected_n = first(n_picks),
    candidates_n = .N
  ), by = stratum]
  if (checks[, any(selected_n != expected_n)]) {
    stop("At least one selector query does not contain exactly k selections")
  }
  if (checks[, any(selected_n <= 0L | selected_n >= candidates_n)]) {
    stop("At least one selector query lacks outcome variation")
  }
  stacked
}

fit_stacked <- function(choice_set, features, formula, draw) {
  stacked <- build_stacked(choice_set, features, draw)
  cat("Fitting tie draw", draw, "on", format(nrow(stacked), big.mark = ","), "rows\n")
  fit_warnings <- character()
  fit <- withCallingHandlers(
    survival::clogit(
      formula,
      data = stacked,
      method = "efron",
      model = FALSE,
      x = FALSE,
      y = FALSE
    ),
    warning = function(condition) {
      fit_warnings <<- c(fit_warnings, conditionMessage(condition))
      invokeRestart("muffleWarning")
    }
  )
  attr(fit, "fit_warnings") <- unique(fit_warnings)
  if (anyNA(stats::coef(fit))) stop("Model contains unidentified coefficients")
  convergence_warnings <- grep(
    "infinite|did not converge|ran out of iterations",
    fit_warnings,
    value = TRUE,
    ignore.case = TRUE
  )
  if (length(convergence_warnings) && identical(provenance$watermark, "INFERENCE")) {
    stop(
      "Inference model did not converge: ",
      paste(unique(convergence_warnings), collapse = " | ")
    )
  }
  fit
}

selector_summary <- function(fit, features, labels, standardize) {
  covariance <- stats::vcov(fit)
  rows <- lapply(features, function(feature) {
    interaction <- interaction_name(fit, feature)
    beta <- stats::coef(fit)[[feature]]
    gamma <- stats::coef(fit)[[interaction]]
    audience_se <- sqrt(covariance[feature, feature])
    difference_se <- sqrt(covariance[interaction, interaction])
    curator_se <- sqrt(
      covariance[feature, feature] + covariance[interaction, interaction] +
        2 * covariance[feature, interaction]
    )
    interval <- function(estimate, se) {
      c(estimate - stats::qnorm(0.975) * se, estimate + stats::qnorm(0.975) * se)
    }
    audience_ci <- interval(beta, audience_se)
    curator_ci <- interval(beta + gamma, curator_se)
    difference_ci <- interval(gamma, difference_se)
    data.table(
      term = feature,
      feature = labels[[feature]],
      unit = if (standardize[[feature]]) "One standard deviation" else "Absent to present",
      audience_log_odds = unname(beta),
      audience_se = audience_se,
      audience_conf_low = audience_ci[[1L]],
      audience_conf_high = audience_ci[[2L]],
      audience_odds_ratio = exp(beta),
      curator_log_odds = unname(beta + gamma),
      curator_se = curator_se,
      curator_conf_low = curator_ci[[1L]],
      curator_conf_high = curator_ci[[2L]],
      curator_odds_ratio = exp(beta + gamma),
      curator_minus_audience_log_odds = unname(gamma),
      difference_se = difference_se,
      difference_conf_low = difference_ci[[1L]],
      difference_conf_high = difference_ci[[2L]],
      difference_p_value = 2 * stats::pnorm(-abs(gamma / difference_se))
    )
  })
  output <- rbindlist(rows)
  output[, difference_q_value_bh := stats::p.adjust(difference_p_value, method = "BH")]
  output
}

probability_summary <- function(
    fit, features, labels, standardize, article_rates, draws, seed) {
  covariance <- stats::vcov(fit)
  p0 <- article_rates$n_picks / article_rates$n_candidates
  probability_change <- function(beta) {
    mean(stats::plogis(stats::qlogis(p0) + beta) - p0)
  }
  rows <- lapply(seq_along(features), function(index) {
    feature <- features[[index]]
    interaction <- interaction_name(fit, feature)
    names_pair <- c(feature, interaction)
    estimates <- stats::coef(fit)[names_pair]
    covariance_pair <- covariance[names_pair, names_pair, drop = FALSE]
    set.seed(seed + index * 100L)
    simulated <- MASS::mvrnorm(draws, mu = estimates, Sigma = covariance_pair)
    audience_draws <- apply(simulated, 1L, function(x) probability_change(x[[1L]]))
    curator_draws <- apply(simulated, 1L, function(x) probability_change(sum(x)))
    difference_draws <- curator_draws - audience_draws
    interval <- function(x) unname(stats::quantile(x, c(0.025, 0.975)))
    audience_ci <- interval(audience_draws)
    curator_ci <- interval(curator_draws)
    difference_ci <- interval(difference_draws)
    data.table(
      term = feature,
      feature = labels[[feature]],
      unit = if (standardize[[feature]]) "One standard deviation" else "Absent to present",
      reference_probability_mean = mean(p0),
      audience_change_pp = 100 * probability_change(estimates[[1L]]),
      audience_conf_low_pp = 100 * audience_ci[[1L]],
      audience_conf_high_pp = 100 * audience_ci[[2L]],
      curator_change_pp = 100 * probability_change(sum(estimates)),
      curator_conf_low_pp = 100 * curator_ci[[1L]],
      curator_conf_high_pp = 100 * curator_ci[[2L]],
      curator_minus_audience_pp = 100 * (
        probability_change(sum(estimates)) - probability_change(estimates[[1L]])
      ),
      difference_conf_low_pp = 100 * difference_ci[[1L]],
      difference_conf_high_pp = 100 * difference_ci[[2L]]
    )
  })
  rbindlist(rows)
}
```

# Fit one candidate scope

```{r run-scope}
run_scope <- function(scope) {
  cat("\n---", toupper(scope), "CANDIDATE MODEL ---\n")
  input_path <- file.path(feature_root, paste0("choice_set_", scope, ".parquet"))
  if (!file.exists(input_path)) stop("Missing choice set: ", input_path)
  choice_set <- as.data.table(arrow::read_parquet(input_path))
  features <- unlist(feature_manifest$models[[scope]]$features, use.names = FALSE)
  missing <- setdiff(
    c("story_id", "comment_id", "n_candidates", "n_picks", "curator_selected", features),
    names(choice_set)
  )
  if (length(missing)) stop("Missing columns in ", scope, ": ", paste(missing, collapse = ", "))
  if (choice_set[, anyDuplicated(paste(story_id, comment_id))]) {
    stop("Candidate keys are not unique in ", scope)
  }

  labels <- setNames(
    vapply(features, function(feature) feature_manifest$features[[feature]]$label, character(1)),
    features
  )
  standardize <- setNames(
    vapply(features, function(feature) isTRUE(feature_manifest$features[[feature]]$standardize), logical(1)),
    features
  )
  scaling <- rbindlist(lapply(features, function(feature) {
    if (!standardize[[feature]]) {
      return(data.table(term = feature, mean = NA_real_, sd = NA_real_, standardized = FALSE))
    }
    scaled <- safe_scale(choice_set[[feature]], feature)
    set(choice_set, j = feature, value = scaled$value)
    data.table(term = feature, mean = scaled$mean, sd = scaled$sd, standardized = TRUE)
  }))
  if (choice_set[, anyNA(.SD), .SDcols = features]) stop("Missing values remain after scaling")

  right_hand_side <- c(
    features,
    paste0("curator:", features),
    "strata(stratum)",
    "cluster(story_id)"
  )
  model_formula <- stats::as.formula(
    paste("selected ~", paste(right_hand_side, collapse = " + "))
  )
  primary <- fit_stacked(choice_set, features, model_formula, 1L)
  covariance <- stats::vcov(primary)
  coefficients <- data.table(
    term = names(stats::coef(primary)),
    estimate = unname(stats::coef(primary)),
    std_error_article_clustered = sqrt(diag(covariance))
  )
  coefficients[, `:=`(
    conf_low = estimate - stats::qnorm(0.975) * std_error_article_clustered,
    conf_high = estimate + stats::qnorm(0.975) * std_error_article_clustered,
    p_value = 2 * stats::pnorm(-abs(estimate / std_error_article_clustered))
  )]
  associations <- selector_summary(primary, features, labels, standardize)
  article_rates <- unique(choice_set[, .(story_id, n_candidates, n_picks)])
  probabilities <- probability_summary(
    primary, features, labels, standardize, article_rates,
    probability_draws, base_seed
  )

  available_draws <- length(grep("^audience_selected_draw_[0-9]+$", names(choice_set)))
  tie_draws <- min(tie_draws_requested, available_draws)
  sensitivity <- rbindlist(lapply(seq_len(tie_draws), function(draw) {
    fit <- if (draw == 1L) primary else fit_stacked(
      choice_set, features, model_formula, draw
    )
    output <- data.table(
      draw = draw,
      term = features,
      curator_minus_audience_log_odds = vapply(
        features,
        function(feature) unname(stats::coef(fit)[[interaction_name(fit, feature)]]),
        numeric(1)
      )
    )
    if (draw > 1L) rm(fit)
    output
  }))
  sensitivity_summary <- sensitivity[, .(
    draws = .N,
    estimate_median = stats::median(curator_minus_audience_log_odds),
    estimate_min = min(curator_minus_audience_log_odds),
    estimate_max = max(curator_minus_audience_log_odds),
    estimate_sd = if (.N > 1L) stats::sd(curator_minus_audience_log_odds) else 0,
    sign_changes_across_draws = min(curator_minus_audience_log_odds) < 0 &
      max(curator_minus_audience_log_odds) > 0
  ), by = term]
  sensitivity_summary[, feature := labels[term]]

  summary <- data.table(
    statistic = c(
      "candidate_scope", "eligible_articles", "candidate_comments",
      "curator_selections", "stacked_rows", "tie_draws_fitted",
      "article_selector_strata", "watermark"
    ),
    value = c(
      scope,
      uniqueN(choice_set$story_id),
      nrow(choice_set),
      sum(choice_set$curator_selected),
      2L * nrow(choice_set),
      tie_draws,
      2L * uniqueN(choice_set$story_id),
      provenance$watermark
    )
  )

  scope_output <- file.path(output_root, scope)
  dir.create(scope_output, recursive = TRUE, showWarnings = FALSE)
  fwrite(scaling, file.path(scope_output, "feature_scaling.csv"))
  fwrite(coefficients, file.path(scope_output, "model_coefficients.csv"))
  fwrite(associations, file.path(scope_output, "selector_associations.csv"))
  fwrite(probabilities, file.path(scope_output, "probability_contrasts.csv"))
  fwrite(sensitivity, file.path(scope_output, "tie_sensitivity_draws.csv"))
  fwrite(sensitivity_summary, file.path(scope_output, "tie_sensitivity_summary.csv"))
  fwrite(summary, file.path(scope_output, "sample_summary.csv"))
  environment(primary$terms) <- baseenv()
  saveRDS(primary, file.path(scope_output, "stacked_selection_model.rds"))
  writeLines(
    c(
      paste("Candidate scope:", scope),
      paste("Feature watermark:", provenance$watermark),
      paste("Eligible articles:", uniqueN(choice_set$story_id)),
      paste("Candidate comments:", nrow(choice_set)),
      paste("Tie draws fitted:", tie_draws),
      "Article controls: article-by-selector conditional-logit strata.",
      "Uncertainty: standard errors clustered by article.",
      "Audience outcome: top-k final relative-vote selection.",
      "Interpretation: retrospective within-discussion associations, not causal effects.",
      if (length(attr(primary, "fit_warnings"))) {
        paste("Fit warnings:", paste(attr(primary, "fit_warnings"), collapse = " | "))
      } else {
        "Fit warnings: none"
      }
    ),
    file.path(scope_output, "diagnostics.txt")
  )
  list(scope = scope, associations = associations, probabilities = probabilities, summary = summary)
}

results <- lapply(c("root", "all"), run_scope)
```

# Compact results

```{r results, echo=FALSE}
for (result in results) {
  cat("\n##", if (result$scope == "root") "Root candidates" else "All comments", "\n\n")
  print(knitr::kable(result$summary, caption = "Sample accounting"))
  print(knitr::kable(
    result$associations[, .(
      feature,
      audience_log_odds,
      curator_log_odds,
      curator_minus_audience_log_odds,
      difference_conf_low,
      difference_conf_high,
      difference_q_value_bh
    )],
    digits = 4,
    caption = "Selector-specific associations on a common selection scale"
  ))
}
```

# Interpretation limits

The audience label is based on final collection-time vote totals and therefore
reflects exposure, position, social influence, and retrospective survival as
well as preference. Author-history vote variables are also collection-snapshot
reputation measures. Curators include journalists, community-management staff,
and volunteer guides. The models estimate associations only within discussions
where at least one usable sticky selection was observed.
