---
title: "Conditional-logit Efron versus exact sensitivity (2025)"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Objective

Assess whether the Efron approximation used by the primary stacked
conditional-logit models materially changes coefficient signs, magnitudes, or
within-query rankings. This is a development-only sensitivity analysis, not a
replacement model and not a second test-set evaluation.

Exact conditional likelihood becomes expensive as `choose(n_candidates,
n_picks)` grows. The analysis therefore uses a deterministic subset of
development articles whose strata satisfy a prespecified combination limit and
a reduced, substantively representative feature set.

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("arrow", "data.table", "jsonlite", "survival", "knitr")
missing_packages <- required_packages[!vapply(
  required_packages, requireNamespace, logical(1), quietly = TRUE
)]
if (length(missing_packages)) {
  stop("Restore the R environment before knitting. Missing: ",
       paste(missing_packages, collapse = ", "))
}
library(data.table)
library(survival)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Configuration

```{r configuration}
feature_root <- Sys.getenv(
  "COMMENTGAP_MODEL_DATA_ROOT",
  "model_output/selection_2025/model_data"
)

output_root <- Sys.getenv(
  "COMMENTGAP_REGRESSION_SENSITIVITY_ROOT",
  "model_output/selection_2025/regression_sensitivity/efron_exact_shared_preprocessing_v4"
)
sample_articles <- as.integer(Sys.getenv(
  "COMMENTGAP_EXACT_SAMPLE_ARTICLES", "100"
))
max_combinations <- as.numeric(Sys.getenv(
  "COMMENTGAP_EXACT_MAX_COMBINATIONS", "5000"
))
base_seed <- as.integer(Sys.getenv("COMMENTGAP_MODEL_SEED", "20260813"))

if (sample_articles < 25L) {
  stop("COMMENTGAP_EXACT_SAMPLE_ARTICLES must be at least 25")
}
if (!is.finite(max_combinations) || max_combinations < 100) {
  stop("COMMENTGAP_EXACT_MAX_COMBINATIONS must be at least 100")
}
dir.create(output_root, recursive = TRUE, showWarnings = FALSE)

split_path <- file.path(feature_root, "master_article_split.parquet")
manifest_path <- file.path(feature_root, "feature_manifest.json")
if (!file.exists(split_path)) stop("Missing shared split: ", split_path)
parameters_path <- file.path(feature_root, "preprocessing_parameters.json")
if (!file.exists(manifest_path)) stop("Missing feature manifest: ", manifest_path)
if (!file.exists(parameters_path)) stop("Missing preprocessing parameters: ", parameters_path)

article_split <- as.data.table(arrow::read_parquet(split_path))
article_split[, story_id := as.character(story_id)]
if (article_split[, anyDuplicated(story_id)]) {
  stop("Shared split contains duplicate articles")
}
if (!setequal(unique(article_split$split_role),
              c("development", "paper2_test"))) {
  stop("Shared split roles are invalid")
}
if (article_split[split_role == "paper2_test",
                  any(development_fold != -1L)]) {
  stop("Paper 2 test articles must not have development folds")
}
development_ids <- article_split[
  split_role == "development", unique(story_id)
]
if (!length(development_ids)) stop("Shared split has no development articles")

feature_manifest <- jsonlite::fromJSON(
  manifest_path, simplifyVector = FALSE
)
preprocessing_parameters <- jsonlite::fromJSON(parameters_path, simplifyVector = FALSE)
if (!identical(as.integer(preprocessing_parameters$version), 4L)) {
  stop("Expected shared preprocessing version 4; force-rerun 06A2")
}
```

# Prespecified reduced specification

The reduced specification retains text form, affect/toxicity, article
relationship, timing/order, discussion pace, recent activity share, prior reply composition,
author history, and four AQuA dimensions. The all-comment model additionally
retains reply status and development-mean-centred reply depth. Scope-specific
novelty variables are used. Every sensitivity feature must belong to the
corresponding primary production model.

```{r feature-contract}
common_features <- c(
  "log_words",
  "sentiment_positive",
  "sentiment_negative",
  "toxicity_probability",
  "lexdiv_length_adjusted",
  "url_present",
  "article_similarity_top3",
  "log_hours_since_article",
  "prior_reply_composition",
  "discussion_pace",
  "log_author_prior_30d_comments",
  "author_prior_30d_upvote_reception",
  "aqua_relevance_expected",
  "aqua_fact_expected",
  "aqua_justification_expected",
  "aqua_respect_expected"
)
features_by_scope <- list(
  root = c(common_features, "novelty_prior_roots_model"),
  all = c(
    common_features,
    "novelty_prior_all_model",
    "is_reply",
    "reply_depth_centered"
  )
)
retired_features <- c("log_prior_roots", "log_depth")
for (scope in names(features_by_scope)) {
  production_features <- unlist(
    feature_manifest$models[[scope]]$features,
    use.names = FALSE
  )
  retired <- intersect(features_by_scope[[scope]], retired_features)
  if (length(retired)) {
    stop(scope, " sensitivity specification contains retired features: ",
         paste(retired, collapse = ", "))
  }
  missing <- setdiff(features_by_scope[[scope]], production_features)
  if (length(missing)) {
    stop(scope, " reduced sensitivity features are absent from the ",
         "production contract: ", paste(missing, collapse = ", "))
  }
}
```

# Helpers

```{r helpers}
safe_scale <- function(x, name) {
  x <- as.numeric(x)
  spread <- stats::sd(x)
  if (!is.finite(spread) || spread <= 0) {
    stop("Cannot standardize feature with no variation: ", name)
  }
  list(value = (x - mean(x)) / spread, mean = mean(x), sd = spread)
}

build_stacked <- function(choice_set, features) {
  keep <- c(
    "story_id", "comment_id", "n_candidates", "n_picks", features
  )
  audience <- copy(choice_set[, ..keep])
  audience[, selector := "audience"]
  audience[, curator := 0L]
  audience[, selected := as.integer(
    choice_set$audience_selected_draw_01
  )]

  curator_rows <- copy(choice_set[, ..keep])
  curator_rows[, selector := "curator"]
  curator_rows[, curator := 1L]
  curator_rows[, selected := as.integer(choice_set$curator_selected)]

  stacked <- rbindlist(list(audience, curator_rows), use.names = TRUE)
  stacked[, stratum := paste(story_id, selector, sep = "::")]
  setorder(stacked, stratum, comment_id)
  checks <- stacked[, .(
    selected_n = sum(selected),
    expected_n = first(n_picks),
    candidates_n = .N
  ), by = stratum]
  if (checks[, any(selected_n != expected_n)]) {
    stop("A sensitivity stratum does not contain exactly k selections")
  }
  if (checks[, any(selected_n <= 0L | selected_n >= candidates_n)]) {
    stop("A sensitivity stratum lacks outcome variation")
  }
  stacked
}

fit_method <- function(stacked, formula, method) {
  warnings <- character()
  elapsed <- system.time({
    fit <- withCallingHandlers(
      survival::clogit(
        formula,
        data = stacked,
        method = method,
        model = TRUE,
        x = TRUE,
        y = TRUE
      ),
      warning = function(condition) {
        warnings <<- c(warnings, conditionMessage(condition))
        invokeRestart("muffleWarning")
      }
    )
  })
  if (anyNA(stats::coef(fit))) {
    stop(method, " sensitivity fit contains unidentified coefficients")
  }
  list(
    fit = fit,
    warnings = unique(warnings),
    elapsed_seconds = unname(elapsed[["elapsed"]])
  )
}

coefficient_comparison <- function(efron, exact) {
  efron_coef <- stats::coef(efron)
  exact_coef <- stats::coef(exact)
  if (!identical(sort(names(efron_coef)), sort(names(exact_coef)))) {
    stop("Efron and exact fits expose different coefficient sets")
  }
  terms <- names(efron_coef)
  output <- data.table(
    term = terms,
    efron = unname(efron_coef[terms]),
    exact = unname(exact_coef[terms])
  )
  output[, difference_exact_minus_efron := exact - efron]
  output[, absolute_difference := abs(difference_exact_minus_efron)]
  output[, sign_agreement := sign(efron) == sign(exact)]
  output[, material_sign_flip := !sign_agreement &
    pmax(abs(efron), abs(exact)) >= 0.10]
  output
}
```

# Run one candidate scope

```{r run-scope}
run_scope <- function(scope) {
  cat("\n---", toupper(scope), "EFRON/EXACT SENSITIVITY ---\n")
  features <- features_by_scope[[scope]]
  input_path <- file.path(
    feature_root, paste0("choice_set_", scope, ".parquet")
  )
  if (!file.exists(input_path)) stop("Missing choice set: ", input_path)
  choice_set <- as.data.table(arrow::read_parquet(input_path))
  choice_set[, story_id := as.character(story_id)]
  choice_set <- choice_set[story_id %in% development_ids]

  required <- c(
    "story_id", "comment_id", "article_month", "n_candidates",
    "n_picks", "curator_selected", "audience_selected_draw_01",
    features
  )
  missing <- setdiff(required, names(choice_set))
  if (length(missing)) {
    stop(scope, " sensitivity input is missing: ",
         paste(missing, collapse = ", "))
  }

  article <- unique(choice_set[, .(
    story_id, article_month, n_candidates, n_picks
  )])
  article[, log_combinations := lchoose(n_candidates, n_picks)]
  article[, exact_eligible :=
    log_combinations <= log(max_combinations)]
  eligible <- article[exact_eligible == TRUE]
  if (nrow(eligible) < 25L) {
    stop(scope, " has too few exact-likelihood-eligible articles")
  }

  eligible[, size_tercile := pmin(
    2L,
    as.integer(3 * (frank(
      n_candidates, ties.method = "average"
    ) - 1) / .N)
  )]
  set.seed(base_seed + if (scope == "root") 101L else 202L)
  eligible[, random_order := runif(.N)]
  setorder(eligible, article_month, size_tercile, random_order, story_id)

  # Round-robin through month-by-size cells to retain broad coverage.
  eligible[, cell_position := seq_len(.N),
           by = .(article_month, size_tercile)]
  setorder(eligible, cell_position, random_order, story_id)
  sampled_articles <- eligible[
    seq_len(min(sample_articles, .N))
  ]
  sampled_ids <- sampled_articles$story_id
  sensitivity <- choice_set[story_id %in% sampled_ids]

  scaling <- rbindlist(lapply(features, function(feature) {
    if (!isTRUE(feature_manifest$features[[feature]]$standardize)) {
      return(data.table(
        term = feature, mean = NA_real_, sd = NA_real_,
        standardized = FALSE
      ))
    }
    scaled <- safe_scale(sensitivity[[feature]], feature)
    set(sensitivity, j = feature, value = scaled$value)
    data.table(
      term = feature, mean = scaled$mean, sd = scaled$sd,
      standardized = TRUE
    )
  }))
  if (sensitivity[, anyNA(.SD), .SDcols = features]) {
    stop("Missing values remain in sensitivity features")
  }

  stacked <- build_stacked(sensitivity, features)
  formula <- stats::as.formula(paste(
    "selected ~",
    paste(c(
      features,
      paste0("curator:", features),
      "strata(stratum)"
    ), collapse = " + ")
  ))

  efron_result <- fit_method(stacked, formula, "efron")
  exact_result <- fit_method(stacked, formula, "exact")
  comparison <- coefficient_comparison(
    efron_result$fit, exact_result$fit
  )

  efron_lp <- as.numeric(stats::predict(
    efron_result$fit, type = "lp", reference = "zero"
  ))
  exact_lp <- as.numeric(stats::predict(
    exact_result$fit, type = "lp", reference = "zero"
  ))
  prediction_agreement <- copy(stacked[, .(
    story_id, comment_id, selector, stratum
  )])
  prediction_agreement[, efron_score := efron_lp]
  prediction_agreement[, exact_score := exact_lp]
  query_agreement <- prediction_agreement[, .(
    spearman = suppressWarnings(stats::cor(
      efron_score, exact_score,
      method = "spearman", use = "complete.obs"
    )),
    top_comment_agreement = (
      comment_id[which.max(efron_score)] ==
      comment_id[which.max(exact_score)]
    )
  ), by = .(story_id, selector)]

  summary <- data.table(
    statistic = c(
      "scope",
      "development_articles_available",
      "exact_eligible_articles",
      "sampled_articles",
      "candidate_rows",
      "stacked_rows",
      "maximum_choose_n_k",
      "reduced_features",
      "coefficients",
      "coefficient_sign_agreement",
      "material_sign_flips",
      "coefficient_spearman",
      "median_absolute_coefficient_difference",
      "maximum_absolute_coefficient_difference",
      "median_query_score_spearman",
      "top_comment_agreement",
      "efron_elapsed_seconds",
      "exact_elapsed_seconds",
      "efron_warnings",
      "exact_warnings"
    ),
    value = c(
      scope,
      nrow(article),
      nrow(eligible),
      nrow(sampled_articles),
      nrow(sensitivity),
      nrow(stacked),
      max(exp(sampled_articles$log_combinations)),
      length(features),
      nrow(comparison),
      mean(comparison$sign_agreement),
      sum(comparison$material_sign_flip),
      stats::cor(
        comparison$efron, comparison$exact,
        method = "spearman"
      ),
      stats::median(comparison$absolute_difference),
      max(comparison$absolute_difference),
      stats::median(query_agreement$spearman, na.rm = TRUE),
      mean(query_agreement$top_comment_agreement),
      efron_result$elapsed_seconds,
      exact_result$elapsed_seconds,
      paste(efron_result$warnings, collapse = " | "),
      paste(exact_result$warnings, collapse = " | ")
    )
  )

  scope_output <- file.path(output_root, scope)
  dir.create(scope_output, recursive = TRUE, showWarnings = FALSE)
  fwrite(scaling, file.path(scope_output, "feature_scaling.csv"))
  fwrite(sampled_articles, file.path(
    scope_output, "sampled_articles.csv"
  ))
  jsonlite::write_json(list(
    analysis = "development-only Efron versus exact sensitivity",
    scope = scope,
    created_at = format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z"),
    shared_preprocessing_version = as.integer(preprocessing_parameters$version),
    feature_manifest_path = normalizePath(manifest_path, mustWork = TRUE),
    feature_manifest_md5 = unname(tools::md5sum(manifest_path)),
    development_articles_available = length(development_ids),
    sampled_articles = nrow(sampled_articles),
    features = features,
    sealed_test_performance_read = FALSE
  ), file.path(scope_output, "analysis_manifest.json"), auto_unbox = TRUE, pretty = TRUE)
  fwrite(comparison, file.path(
    scope_output, "coefficient_comparison.csv"
  ))
  fwrite(query_agreement, file.path(
    scope_output, "query_ranking_agreement.csv"
  ))
  fwrite(summary, file.path(scope_output, "sensitivity_summary.csv"))
  saveRDS(efron_result$fit, file.path(
    scope_output, "efron_reduced_model.rds"
  ))
  saveRDS(exact_result$fit, file.path(
    scope_output, "exact_reduced_model.rds"
  ))

  list(scope = scope, summary = summary, comparison = comparison)
}

results <- lapply(c("root", "all"), run_scope)
```

# Results

```{r results, echo=FALSE}
for (result in results) {
  cat("\n##", if (result$scope == "root") {
    "Root candidates"
  } else {
    "All comments"
  }, "\n\n")
  print(knitr::kable(
    result$summary,
    caption = "Efron versus exact sensitivity summary"
  ))
  print(knitr::kable(
    head(result$comparison[order(-absolute_difference)], 15),
    digits = 4,
    caption = "Largest coefficient differences"
  ))
}
```

# Interpretation

This analysis addresses approximation sensitivity only. Strong evidence of
robustness consists of high coefficient and within-query score correlations,
few material sign reversals, and high top-comment agreement. Differences do not
invalidate the primary Efron model automatically: the exact analysis uses a
smaller, lower-complexity article sample and a reduced specification.

Neither fit uses Paper 2 test articles. Standard errors are not compared because
article clustering changes uncertainty estimates but not point estimates, while
the purpose here is to isolate the likelihood approximation.
