---
title: "7. Stacked curator–audience selection models (2025)"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Objective

Fit a common-scale conditional-logit model for the requested candidate scopes.
All published comments are the Paper 1 primary scope; root comments are an
opt-in appendix sensitivity. Each comment is stacked once for the audience
selector and once for the curator selector.
Article-by-selector strata provide the article controls; standard errors are
clustered by article. Each specification includes the 20 continuous AQuA
expected ordinal dimensions and excludes the derived AQuA composite.
Models and feature scaling are fitted on the same 50% development articles used
by stage 8; the other 50% are used only for frozen held-out evaluation. The stored
split role retains the legacy name `paper2_test`.

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "arrow", "data.table", "digest", "jsonlite", "survival", "MASS", "knitr"
)
missing_packages <- required_packages[!vapply(
  required_packages, requireNamespace, logical(1), quietly = TRUE
)]
if (length(missing_packages)) {
  stop(
    "Restore the R environment before knitting. Missing: ",
    paste(missing_packages, collapse = ", ")
  )
}
library(data.table)
library(survival)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Configuration and shared feature contract

```{r configuration}
source_feature_root <- Sys.getenv(
  "COMMENTGAP_FEATURE_ROOT",
  "model_output/selection_2025/features"
)
feature_root <- Sys.getenv(
  "COMMENTGAP_MODEL_DATA_ROOT",
  "model_output/selection_2025/model_data"
)
output_root <- Sys.getenv(
  "COMMENTGAP_REGRESSION_ROOT",
  "model_output/selection_2025/regression"
)
split_path <- file.path(feature_root, "master_article_split.parquet")
tie_draws_requested <- as.integer(Sys.getenv("COMMENTGAP_TIE_DRAWS", "10"))
probability_draws <- as.integer(Sys.getenv("COMMENTGAP_PROBABILITY_DRAWS", "5000"))
test_bootstrap_draws <- as.integer(Sys.getenv("COMMENTGAP_TEST_BOOTSTRAP_DRAWS", "1000"))
base_seed <- as.integer(Sys.getenv("COMMENTGAP_MODEL_SEED", "20260813"))
fit_cache_root <- Sys.getenv(
  "COMMENTGAP_REGRESSION_CACHE_ROOT",
  file.path(output_root, "fit_cache")
)
log_root <- Sys.getenv("COMMENTGAP_LOG_ROOT", "logs")
force_recompute <- tolower(Sys.getenv(
  "COMMENTGAP_REGRESSION_FORCE_RECOMPUTE", "0"
)) %in% c("1", "true", "yes")
requested_scopes <- strsplit(
  Sys.getenv(
    "COMMENTGAP_REGRESSION_SCOPES",
    Sys.getenv("COMMENTGAP_MODEL_SCOPES", "all")
  ),
  ",",
  fixed = TRUE
)[[1L]]
requested_scopes <- unique(trimws(requested_scopes))
requested_scopes <- requested_scopes[nzchar(requested_scopes)]
fit_cache_schema <- "stacked-clogit-efron-shared-preprocessing-v4"

if (tie_draws_requested < 1L) stop("COMMENTGAP_TIE_DRAWS must be positive")
if (probability_draws < 100L) stop("COMMENTGAP_PROBABILITY_DRAWS must be at least 100")
if (test_bootstrap_draws < 100L) {
  stop("COMMENTGAP_TEST_BOOTSTRAP_DRAWS must be at least 100")
}
if (!length(requested_scopes) ||
    !all(requested_scopes %in% c("all", "root"))) {
  stop("COMMENTGAP_REGRESSION_SCOPES must contain all and/or root")
}
if (!file.exists(split_path)) {
  stop("Run 03_shared_model_preprocessing.ipynb first; missing shared split: ", split_path)
}
dir.create(output_root, recursive = TRUE, showWarnings = FALSE)
dir.create(fit_cache_root, recursive = TRUE, showWarnings = FALSE)
progress_log_path <- Sys.getenv(
  "COMMENTGAP_REGRESSION_PROGRESS_LOG",
  file.path(log_root, "07_stacked_selection_models_progress.log")
)
dir.create(dirname(progress_log_path), recursive = TRUE, showWarnings = FALSE)
progress_state <- new.env(parent = emptyenv())
progress_state$run_started <- Sys.time()
progress_state$completed_fits <- 0L
progress_state$total_fits <- length(requested_scopes) * tie_draws_requested
progress_state$fit_seconds <- numeric()

format_duration <- function(seconds) {
  if (!is.finite(seconds)) return("unknown")
  seconds <- max(0, round(seconds))
  sprintf(
    "%02d:%02d:%02d",
    seconds %/% 3600L,
    (seconds %% 3600L) %/% 60L,
    seconds %% 60L
  )
}

progress_message <- function(message) {
  elapsed <- as.numeric(difftime(
    Sys.time(), progress_state$run_started, units = "secs"
  ))
  line <- sprintf(
    "[%s] elapsed=%s | %s",
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"),
    format_duration(elapsed),
    message
  )
  cat(line, "\n")
  cat(line, "\n", file = progress_log_path, append = TRUE)
  flush.console()
}

progress_message(sprintf(
  paste0(
    "Run started; scopes=%s; planned conditional-logit fits=%d ",
    "(%d scopes x %d tie draws); cache=%s; force_recompute=%s"
  ),
  paste(requested_scopes, collapse = ","),
  progress_state$total_fits, length(requested_scopes), tie_draws_requested,
  normalizePath(fit_cache_root, mustWork = FALSE), force_recompute
))

article_split <- as.data.table(arrow::read_parquet(split_path))
required_split <- c("story_id", "split_role", "development_fold", "split_stratum")
missing_split <- setdiff(required_split, names(article_split))
if (length(missing_split)) stop("Shared split is missing: ", paste(missing_split, collapse = ", "))
article_split[, story_id := as.character(story_id)]
if (article_split[, anyDuplicated(story_id)]) stop("Shared split contains duplicate articles")
if (!setequal(unique(article_split$split_role), c("development", "paper2_test"))) {
  stop("Shared split must contain development and paper2_test roles")
}

manifest_path <- file.path(feature_root, "feature_manifest.json")
parameters_path <- file.path(feature_root, "preprocessing_parameters.json")
provenance_path <- file.path(source_feature_root, "provenance_manifest.json")
if (!file.exists(manifest_path)) stop("Missing feature manifest: ", manifest_path)
if (!file.exists(parameters_path)) stop("Missing preprocessing parameters: ", parameters_path)
if (!file.exists(provenance_path)) stop("Missing provenance manifest: ", provenance_path)
preprocessing_parameters <- jsonlite::fromJSON(parameters_path, simplifyVector = FALSE)
if (!identical(as.integer(preprocessing_parameters$version), 4L)) stop("Expected shared preprocessing version 4; force-rerun stage 3")
feature_manifest <- jsonlite::fromJSON(manifest_path, simplifyVector = FALSE)
provenance <- jsonlite::fromJSON(provenance_path, simplifyVector = FALSE)
aqua_expected_features <- setdiff(
  grep("^aqua_.*_expected$", names(feature_manifest$features), value = TRUE),
  "aqua_score_expected"
)
if (length(aqua_expected_features) != 20L) {
  stop(
    "The primary model contract must expose exactly 20 AQuA expected dimensions; found ",
    length(aqua_expected_features)
  )
}
cat("Feature watermark:", provenance$watermark, "\n")
cat("Primary AQuA expected dimensions:", length(aqua_expected_features), "\n")
```

# Model helpers

```{r helpers}
safe_scale <- function(x, name) {
  x <- as.numeric(x)
  center <- mean(x)
  spread <- stats::sd(x)
  if (!is.finite(spread) || spread <= 0) {
    stop("Cannot standardize feature with no variation: ", name)
  }
  list(value = (x - center) / spread, mean = center, sd = spread)
}

file_fingerprint <- function(path) {
  information <- file.info(path)
  if (!nrow(information) || is.na(information$size)) {
    stop("Cannot fingerprint missing file: ", path)
  }
  list(
    path = normalizePath(path, mustWork = TRUE),
    size = unname(information$size),
    md5 = unname(tools::md5sum(path))
  )
}

atomic_save_rds <- function(object, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  temporary <- tempfile(pattern = "model-", tmpdir = dirname(path))
  on.exit(unlink(temporary), add = TRUE)
  saveRDS(object, temporary, compress = "gzip")
  if (!file.rename(temporary, path)) {
    stop("Could not atomically publish cached model: ", path)
  }
}

coef.commentgap_cached_clogit <- function(object, ...) {
  object$coefficients
}

atomic_write_json <- function(object, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  temporary <- tempfile(pattern = "manifest-", tmpdir = dirname(path))
  on.exit(unlink(temporary), add = TRUE)
  writeLines(
    jsonlite::toJSON(object, auto_unbox = TRUE, pretty = TRUE, null = "null"),
    temporary
  )
  if (!file.rename(temporary, path)) {
    stop("Could not atomically publish cache manifest: ", path)
  }
}

record_completed_fit <- function(scope, draw, fit_seconds = NULL, cached = FALSE) {
  progress_state$completed_fits <- progress_state$completed_fits + 1L
  if (!is.null(fit_seconds)) {
    progress_state$fit_seconds <- c(progress_state$fit_seconds, fit_seconds)
  }
  remaining_fits <- progress_state$total_fits - progress_state$completed_fits
  rough_eta_seconds <- if (length(progress_state$fit_seconds)) {
    mean(progress_state$fit_seconds) * remaining_fits
  } else {
    NA_real_
  }
  progress_message(sprintf(
    paste0(
      "%s scope=%s tie_draw=%d completed=%d/%d rough_ETA=%s ",
      "(fit-average; improves after first all-comments fit)"
    ),
    if (cached) "CACHE HIT" else "DONE",
    scope, draw, progress_state$completed_fits, progress_state$total_fits,
    format_duration(rough_eta_seconds)
  ))
}

interaction_name <- function(fit, feature) {
  candidates <- c(paste0(feature, ":curator"), paste0("curator:", feature))
  found <- candidates[candidates %in% names(stats::coef(fit))]
  if (length(found) != 1L) stop("Cannot identify interaction for ", feature)
  found[[1L]]
}

build_stacked <- function(choice_set, features, draw) {
  label <- sprintf("audience_selected_draw_%02d", draw)
  if (!label %in% names(choice_set)) stop("Missing tie-draw label: ", label)
  keep <- c("story_id", "comment_id", "n_candidates", "n_picks", features)

  audience <- copy(choice_set[, ..keep])
  audience[, `:=`(
    selector = "audience",
    curator = 0L,
    selected = as.integer(choice_set[[label]])
  )]
  curator_rows <- copy(choice_set[, ..keep])
  curator_rows[, `:=`(
    selector = "curator",
    curator = 1L,
    selected = as.integer(choice_set$curator_selected)
  )]
  stacked <- rbindlist(list(audience, curator_rows), use.names = TRUE)
  stacked[, stratum := paste(story_id, selector, sep = "::")]
  setorder(stacked, stratum, comment_id)

  checks <- stacked[, .(
    selected_n = sum(selected),
    expected_n = first(n_picks),
    candidates_n = .N
  ), by = stratum]
  if (checks[, any(selected_n != expected_n)]) {
    stop("At least one selector query does not contain exactly k selections")
  }
  if (checks[, any(selected_n <= 0L | selected_n >= candidates_n)]) {
    stop("At least one selector query lacks outcome variation")
  }
  stacked
}

fit_stacked <- function(
    choice_set, features, formula, draw, scope, cache_fingerprint) {
  cache_directory <- file.path(
    fit_cache_root, scope, sprintf("tie_draw_%02d", draw)
  )
  model_path <- file.path(cache_directory, "model.rds")
  cache_manifest_path <- file.path(cache_directory, "cache_manifest.json")
  if (!force_recompute && file.exists(model_path) && file.exists(cache_manifest_path)) {
    cache_manifest <- tryCatch(
      jsonlite::fromJSON(cache_manifest_path, simplifyVector = TRUE),
      error = function(condition) NULL
    )
    if (!is.null(cache_manifest) &&
        identical(cache_manifest$fingerprint, cache_fingerprint)) {
      cached_fit <- tryCatch(readRDS(model_path), error = function(condition) NULL)
      if (!is.null(cached_fit) &&
          length(stats::coef(cached_fit)) == 2L * length(features) &&
          all(is.finite(stats::coef(cached_fit)))) {
        cached_seconds <- suppressWarnings(as.numeric(cache_manifest$fit_seconds))
        record_completed_fit(
          scope, draw,
          fit_seconds = if (length(cached_seconds) && is.finite(cached_seconds)) {
            cached_seconds
          } else {
            NULL
          },
          cached = TRUE
        )
        return(cached_fit)
      }
      progress_message(sprintf(
        "CACHE INVALID scope=%s tie_draw=%d; cached model could not be validated",
        scope, draw
      ))
    } else {
      progress_message(sprintf(
        "CACHE MISS scope=%s tie_draw=%d; fingerprint changed", scope, draw
      ))
    }
  }
  stacked <- build_stacked(choice_set, features, draw)
  fit_started <- Sys.time()
  progress_message(sprintf(
    "START scope=%s tie_draw=%d rows=%s fit=%d/%d",
    scope, draw, format(nrow(stacked), big.mark = ","),
    progress_state$completed_fits + 1L, progress_state$total_fits
  ))
  fit_warnings <- character()
  fit <- withCallingHandlers(
    survival::clogit(
      formula,
      data = stacked,
      method = "efron",
      model = FALSE,
      x = FALSE,
      y = FALSE
    ),
    warning = function(condition) {
      fit_warnings <<- c(fit_warnings, conditionMessage(condition))
      invokeRestart("muffleWarning")
    }
  )
  attr(fit, "fit_warnings") <- unique(fit_warnings)
  if (anyNA(stats::coef(fit))) stop("Model contains unidentified coefficients")
  convergence_warnings <- grep(
    "infinite|did not converge|ran out of iterations",
    fit_warnings,
    value = TRUE,
    ignore.case = TRUE
  )
  if (length(convergence_warnings) && identical(provenance$watermark, "INFERENCE")) {
    stop(
      "Inference model did not converge: ",
      paste(unique(convergence_warnings), collapse = " | ")
    )
  }
  fit_seconds <- as.numeric(difftime(Sys.time(), fit_started, units = "secs"))
  environment(fit$terms) <- baseenv()
  cache_object <- if (draw == 1L) {
    fit
  } else {
    structure(
      list(
        coefficients = stats::coef(fit),
        fit_warnings = attr(fit, "fit_warnings")
      ),
      class = "commentgap_cached_clogit"
    )
  }
  artifact_kind <- if (draw == 1L) "full_primary_fit" else "tie_draw_coefficients"
  atomic_save_rds(cache_object, model_path)
  atomic_write_json(
    list(
      fingerprint = cache_fingerprint,
      cache_schema = fit_cache_schema,
      scope = scope,
      tie_draw = draw,
      artifact_kind = artifact_kind,
      stacked_rows = nrow(stacked),
      fitted_at = format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z"),
      fit_seconds = fit_seconds,
      r_version = R.version.string,
      survival_version = as.character(utils::packageVersion("survival"))
    ),
    cache_manifest_path
  )
  progress_message(sprintf(
    "CACHE WRITE scope=%s tie_draw=%d fit_time=%s path=%s",
    scope, draw, format_duration(fit_seconds), model_path
  ))
  record_completed_fit(scope, draw, fit_seconds = fit_seconds)
  fit
}


score_frozen_model <- function(fit, choice_set, features, audience_draw = 1L) {
  audience_label <- sprintf("audience_selected_draw_%02d", audience_draw)
  if (!audience_label %in% names(choice_set)) {
    stop("Missing audience tie-draw label: ", audience_label)
  }
  audience_beta <- vapply(features, function(feature) unname(stats::coef(fit)[[feature]]), numeric(1))
  interaction_beta <- vapply(features, function(feature) {
    unname(stats::coef(fit)[[interaction_name(fit, feature)]])
  }, numeric(1))
  matrix_x <- as.matrix(choice_set[, ..features])
  audience_score <- drop(matrix_x %*% audience_beta)
  curator_score <- drop(matrix_x %*% (audience_beta + interaction_beta))
  rbindlist(list(
    choice_set[, .(story_id, comment_id, n_picks, selector = "audience",
      selected = as.integer(get(audience_label)), score = audience_score)],
    choice_set[, .(story_id, comment_id, n_picks, selector = "curator",
      selected = as.integer(curator_selected), score = curator_score)]
  ))
}

evaluate_frozen_scores <- function(scores) {
  scores[, {
    k <- first(n_picks)
    ranked <- order(-score, comment_id)
    ideal <- order(-selected, comment_id)
    predicted <- ranked[seq_len(k)]
    ideal_top <- ideal[seq_len(k)]
    discounts <- 1 / log2(seq_len(k) + 1)
    dcg <- sum(selected[predicted] * discounts)
    idcg <- sum(selected[ideal_top] * discounts)
    actual <- which(selected == 1L)
    overlap <- length(intersect(predicted, actual))
    selected_ranks <- match(actual, ranked)
    .(n_candidates = .N, n_picks = k,
      ndcg_at_k = if (idcg > 0) dcg / idcg else NA_real_,
      top_k_overlap = overlap / k,
      jaccard = overlap / length(union(predicted, actual)),
      mean_selected_rank = mean(selected_ranks))
  }, by = .(story_id, selector)]
}

chance_ranking_metrics <- function(scores) {
  article_structure <- scores[, .(
    n_candidates = .N,
    n_picks = first(n_picks)
  ), by = .(story_id, selector)]
  article_structure[, chance_rate := n_picks / n_candidates]
  article_structure[, expected_jaccard := vapply(
    seq_len(.N),
    function(index) {
      n <- n_candidates[[index]]
      k <- n_picks[[index]]
      overlap <- seq.int(max(0L, 2L * k - n), k)
      probabilities <- stats::dhyper(overlap, k, n - k, k)
      sum(probabilities * overlap / (2L * k - overlap))
    },
    numeric(1)
  )]
  article_structure[, .(
    story_id,
    selector,
    n_candidates,
    n_picks,
    ndcg_at_k = chance_rate,
    top_k_overlap = chance_rate,
    jaccard = expected_jaccard,
    mean_selected_rank = (n_candidates + 1) / 2
  )]
}

bootstrap_metric_summary <- function(article_metrics, draws, seed) {
  metric_names <- c("ndcg_at_k", "top_k_overlap", "jaccard", "mean_selected_rank")
  rbindlist(lapply(seq_along(metric_names), function(metric_index) {
    metric <- metric_names[[metric_index]]
    rbindlist(lapply(c("audience", "curator"), function(selector_name) {
      values <- article_metrics[selector == selector_name][[metric]]
      if (!length(values) || any(!is.finite(values))) {
        stop("Non-finite or missing test metric: ", selector_name, " / ", metric)
      }
      set.seed(seed + metric_index * 100L + match(selector_name, c("audience", "curator")))
      bootstrap_estimates <- replicate(
        draws,
        mean(sample(values, length(values), replace = TRUE))
      )
      interval <- unname(stats::quantile(bootstrap_estimates, c(0.025, 0.975)))
      data.table(
        selector = selector_name,
        metric = metric,
        estimate = mean(values),
        conf_low = interval[[1L]],
        conf_high = interval[[2L]],
        articles = length(values),
        bootstrap_draws = draws
      )
    }))
  }))
}

bootstrap_above_chance <- function(model_metrics, chance_metrics, draws, seed) {
  keys <- c("story_id", "selector")
  model_keys <- model_metrics[, ..keys]
  chance_keys <- chance_metrics[, ..keys]
  setorder(model_keys, story_id, selector)
  setorder(chance_keys, story_id, selector)
  if (anyDuplicated(model_keys) || anyDuplicated(chance_keys) ||
      !identical(model_keys, chance_keys)) {
    stop("Model and chance metrics do not contain identical article-selector keys")
  }
  merged <- merge(
    model_metrics,
    chance_metrics,
    by = keys,
    suffixes = c("_model", "_chance"),
    sort = FALSE
  )
  metric_names <- c("ndcg_at_k", "top_k_overlap", "jaccard", "mean_selected_rank")
  rbindlist(lapply(seq_along(metric_names), function(metric_index) {
    metric <- metric_names[[metric_index]]
    rbindlist(lapply(c("audience", "curator"), function(selector_name) {
      subset <- merged[selector == selector_name]
      differences <- subset[[paste0(metric, "_model")]] -
        subset[[paste0(metric, "_chance")]]
      set.seed(
        seed + metric_index * 100L +
          match(selector_name, c("audience", "curator"))
      )
      bootstrap_estimates <- replicate(
        draws,
        mean(sample(differences, length(differences), replace = TRUE))
      )
      interval <- unname(stats::quantile(
        bootstrap_estimates, c(0.025, 0.975)
      ))
      data.table(
        selector = selector_name,
        metric = metric,
        model_estimate = mean(
          subset[[paste0(metric, "_model")]]
        ),
        chance_estimate = mean(
          subset[[paste0(metric, "_chance")]]
        ),
        estimate_above_chance = mean(differences),
        conf_low = interval[[1L]],
        conf_high = interval[[2L]],
        articles = length(differences),
        bootstrap_draws = draws
      )
    }))
  }))
}

model_diagnostics <- function(fit) {
  fit_summary <- summary(fit)
  concordance <- fit_summary$concordance
  covariance <- stats::vcov(fit)
  eigenvalues <- eigen(covariance, symmetric = TRUE, only.values = TRUE)$values
  positive_eigenvalues <- eigenvalues[eigenvalues > .Machine$double.eps]
  data.table(
    statistic = c(
      "iterations", "null_log_likelihood", "fitted_log_likelihood",
      "likelihood_ratio_chisq", "likelihood_ratio_df", "likelihood_ratio_p_value",
      "concordance", "concordance_se", "maximum_absolute_coefficient",
      "maximum_clustered_standard_error", "covariance_condition_number",
      "fit_warning_count", "all_coefficients_finite", "all_standard_errors_finite"
    ),
    value = c(
      paste(fit$iter, collapse = ","),
      fit$loglik[[1L]],
      fit$loglik[[2L]],
      fit_summary$logtest[["test"]],
      fit_summary$logtest[["df"]],
      fit_summary$logtest[["pvalue"]],
      concordance[["C"]],
      concordance[["se(C)"]],
      max(abs(stats::coef(fit))),
      max(sqrt(diag(covariance))),
      if (length(positive_eigenvalues)) {
        max(positive_eigenvalues) / min(positive_eigenvalues)
      } else {
        Inf
      },
      length(attr(fit, "fit_warnings")),
      all(is.finite(stats::coef(fit))),
      all(is.finite(sqrt(diag(covariance))))
    )
  )
}

selector_summary <- function(fit, features, labels, standardize) {
  covariance <- stats::vcov(fit)
  rows <- lapply(features, function(feature) {
    interaction <- interaction_name(fit, feature)
    beta <- stats::coef(fit)[[feature]]
    gamma <- stats::coef(fit)[[interaction]]
    audience_se <- sqrt(covariance[feature, feature])
    difference_se <- sqrt(covariance[interaction, interaction])
    curator_se <- sqrt(
      covariance[feature, feature] + covariance[interaction, interaction] +
        2 * covariance[feature, interaction]
    )
    interval <- function(estimate, se) {
      c(estimate - stats::qnorm(0.975) * se, estimate + stats::qnorm(0.975) * se)
    }
    audience_ci <- interval(beta, audience_se)
    curator_ci <- interval(beta + gamma, curator_se)
    difference_ci <- interval(gamma, difference_se)
    data.table(
      term = feature,
      feature = labels[[feature]],
      unit = if (standardize[[feature]]) "One standard deviation" else "Absent to present",
      audience_log_odds = unname(beta),
      audience_se = audience_se,
      audience_conf_low = audience_ci[[1L]],
      audience_conf_high = audience_ci[[2L]],
      audience_odds_ratio = exp(beta),
      audience_p_value = 2 * stats::pnorm(-abs(beta / audience_se)),
      curator_log_odds = unname(beta + gamma),
      curator_se = curator_se,
      curator_conf_low = curator_ci[[1L]],
      curator_conf_high = curator_ci[[2L]],
      curator_odds_ratio = exp(beta + gamma),
      curator_p_value = 2 * stats::pnorm(-abs((beta + gamma) / curator_se)),
      curator_minus_audience_log_odds = unname(gamma),
      difference_se = difference_se,
      difference_conf_low = difference_ci[[1L]],
      difference_conf_high = difference_ci[[2L]],
      difference_p_value = 2 * stats::pnorm(-abs(gamma / difference_se))
    )
  })
  output <- rbindlist(rows)
  output[, `:=`(
    audience_q_value_bh = stats::p.adjust(audience_p_value, method = "BH"),
    curator_q_value_bh = stats::p.adjust(curator_p_value, method = "BH"),
    difference_q_value_bh = stats::p.adjust(difference_p_value, method = "BH")
  )]
  output
}

probability_summary <- function(
    fit, features, labels, standardize, article_rates, draws, seed) {
  covariance <- stats::vcov(fit)
  p0 <- article_rates$n_picks / article_rates$n_candidates
  probability_change <- function(beta) {
    mean(stats::plogis(stats::qlogis(p0) + beta) - p0)
  }
  rows <- lapply(seq_along(features), function(index) {
    feature <- features[[index]]
    interaction <- interaction_name(fit, feature)
    names_pair <- c(feature, interaction)
    estimates <- stats::coef(fit)[names_pair]
    covariance_pair <- covariance[names_pair, names_pair, drop = FALSE]
    set.seed(seed + index * 100L)
    simulated <- MASS::mvrnorm(draws, mu = estimates, Sigma = covariance_pair)
    audience_draws <- apply(simulated, 1L, function(x) probability_change(x[[1L]]))
    curator_draws <- apply(simulated, 1L, function(x) probability_change(sum(x)))
    difference_draws <- curator_draws - audience_draws
    interval <- function(x) unname(stats::quantile(x, c(0.025, 0.975)))
    audience_ci <- interval(audience_draws)
    curator_ci <- interval(curator_draws)
    difference_ci <- interval(difference_draws)
    data.table(
      term = feature,
      feature = labels[[feature]],
      unit = if (standardize[[feature]]) "One standard deviation" else "Absent to present",
      interpretation = "Heuristic logistic translation; not an exact fixed-k inclusion probability",
      reference_probability_mean = mean(p0),
      audience_change_pp = 100 * probability_change(estimates[[1L]]),
      audience_conf_low_pp = 100 * audience_ci[[1L]],
      audience_conf_high_pp = 100 * audience_ci[[2L]],
      curator_change_pp = 100 * probability_change(sum(estimates)),
      curator_conf_low_pp = 100 * curator_ci[[1L]],
      curator_conf_high_pp = 100 * curator_ci[[2L]],
      curator_minus_audience_pp = 100 * (
        probability_change(sum(estimates)) - probability_change(estimates[[1L]])
      ),
      difference_conf_low_pp = 100 * difference_ci[[1L]],
      difference_conf_high_pp = 100 * difference_ci[[2L]]
    )
  })
  rbindlist(rows)
}
```

# Fit one candidate scope

```{r run-scope}
run_scope <- function(scope) {
  progress_message(paste("Starting candidate scope:", scope))
  input_path <- file.path(feature_root, paste0("choice_set_", scope, ".parquet"))
  if (!file.exists(input_path)) stop("Missing choice set: ", input_path)
  choice_set <- as.data.table(arrow::read_parquet(input_path))
  features <- unlist(feature_manifest$models[[scope]]$features, use.names = FALSE)
  required_author_features <- c(
    "log_author_prior_30d_comments",
    "author_prior_30d_upvote_reception",
    "author_prior_30d_downvote_reception",
    "log_author_prior_comments_story"
  )
  missing_author <- setdiff(required_author_features, features)
  if (length(missing_author)) {
    stop("The shared author-history contract is incomplete: ",
         paste(missing_author, collapse = ", "))
  }
  retired_author_features <- c(
    "log_author_prior_30d_snapshot_upvotes",
    "log_author_prior_30d_snapshot_downvotes"
  )
  if (any(retired_author_features %in% features)) {
    stop("Retired total author-vote features remain in the primary contract")
  }
  required_activity_features <- "discussion_pace"
  missing_activity <- setdiff(required_activity_features, features)
  if (length(missing_activity)) {
    stop("The shared activity contract is incomplete: ",
         paste(missing_activity, collapse = ", "))
  }
  retired_activity_features <- c(
    "log_prior_comments", "log_comments_prev_hour", "recent_activity_share",
    "log_branch_prior_comments", "log_branch_comments_prev_hour"
  )
  if (any(retired_activity_features %in% features)) {
    stop("Retired activity-count features remain in the primary contract")
  }
  missing_aqua <- setdiff(aqua_expected_features, features)
  if (length(missing_aqua)) {
    stop(
      "The ", scope, " primary model manifest omits AQuA expected dimensions: ",
      paste(missing_aqua, collapse = ", "),
      ". Re-run commentgap-features with the production AQuA store."
    )
  }
  if ("aqua_score_expected" %in% features) {
    stop("The derived AQuA composite must not accompany its 20 constituent dimensions")
  }
  missing <- setdiff(
    c("story_id", "comment_id", "n_candidates", "n_picks", "curator_selected", features),
    names(choice_set)
  )
  if (length(missing)) stop("Missing columns in ", scope, ": ", paste(missing, collapse = ", "))
  if (choice_set[, anyDuplicated(paste(story_id, comment_id))]) {
    stop("Candidate keys are not unique in ", scope)
  }
  choice_set[, story_id := as.character(story_id)]
  choice_set <- merge(
    choice_set,
    article_split[, .(story_id, split_role, development_fold, split_stratum)],
    by = "story_id", all = FALSE, sort = FALSE
  )
  if (uniqueN(choice_set$story_id) != nrow(article_split)) {
    stop(scope, " choice set does not contain every article in the shared split")
  }
  development <- choice_set[split_role == "development"]
  test <- choice_set[split_role == "paper2_test"]
  if (!nrow(development) || !nrow(test)) stop("Development and test rows are both required")

  labels <- setNames(
    vapply(features, function(feature) feature_manifest$features[[feature]]$label, character(1)),
    features
  )
  standardize <- setNames(
    vapply(features, function(feature) isTRUE(feature_manifest$features[[feature]]$standardize), logical(1)),
    features
  )
  scaling <- rbindlist(lapply(features, function(feature) {
    if (!standardize[[feature]]) {
      return(data.table(term = feature, mean = NA_real_, sd = NA_real_, standardized = FALSE))
    }
    scaled <- safe_scale(development[[feature]], feature)
    set(development, j = feature, value = scaled$value)
    set(test, j = feature, value = (as.numeric(test[[feature]]) - scaled$mean) / scaled$sd)
    data.table(term = feature, mean = scaled$mean, sd = scaled$sd, standardized = TRUE)
  }))
  if (development[, anyNA(.SD), .SDcols = features] || test[, anyNA(.SD), .SDcols = features]) {
    stop("Missing values remain after development-fitted scaling")
  }

  right_hand_side <- c(
    features,
    paste0("curator:", features),
    "strata(stratum)",
    "cluster(story_id)"
  )
  model_formula <- stats::as.formula(
    paste("selected ~", paste(right_hand_side, collapse = " + "))
  )
  cache_context <- list(
    cache_schema = fit_cache_schema,
    scope = scope,
    input = file_fingerprint(input_path),
    split = file_fingerprint(split_path),
    feature_manifest = file_fingerprint(manifest_path),
    provenance_manifest = file_fingerprint(provenance_path),
    features = features,
    standardize = standardize,
    scaling = as.data.frame(scaling),
    formula = paste(deparse(model_formula), collapse = " "),
    method = "efron",
    r_version = R.version.string,
    package_versions = vapply(
      c("data.table", "survival"),
      function(package) as.character(utils::packageVersion(package)),
      character(1)
    )
  )
  fit_fingerprint <- function(draw) {
    digest::digest(
      c(cache_context, list(
        tie_draw = draw,
        audience_label = sprintf("audience_selected_draw_%02d", draw)
      )),
      algo = "sha256",
      serialize = TRUE
    )
  }
  primary <- fit_stacked(
    development, features, model_formula, 1L, scope, fit_fingerprint(1L)
  )
  diagnostics <- model_diagnostics(primary)
  covariance <- stats::vcov(primary)
  coefficients <- data.table(
    term = names(stats::coef(primary)),
    estimate = unname(stats::coef(primary)),
    std_error_article_clustered = sqrt(diag(covariance))
  )
  coefficients[, `:=`(
    conf_low = estimate - stats::qnorm(0.975) * std_error_article_clustered,
    conf_high = estimate + stats::qnorm(0.975) * std_error_article_clustered,
    p_value = 2 * stats::pnorm(-abs(estimate / std_error_article_clustered))
  )]
  associations <- selector_summary(primary, features, labels, standardize)
  article_rates <- unique(development[, .(story_id, n_candidates, n_picks)])
  probabilities <- probability_summary(
    primary, features, labels, standardize, article_rates,
    probability_draws, base_seed
  )

  available_draws <- length(grep("^audience_selected_draw_[0-9]+$", names(development)))
  tie_draws <- min(tie_draws_requested, available_draws)
  sensitivity <- rbindlist(lapply(seq_len(tie_draws), function(draw) {
    fit <- if (draw == 1L) primary else fit_stacked(
      development, features, model_formula, draw, scope, fit_fingerprint(draw)
    )
    output <- data.table(
      draw = draw,
      term = features,
      curator_minus_audience_log_odds = vapply(
        features,
        function(feature) unname(stats::coef(fit)[[interaction_name(fit, feature)]]),
        numeric(1)
      )
    )
    if (draw > 1L) rm(fit)
    output
  }))
  sensitivity_summary <- sensitivity[, .(
    draws = .N,
    estimate_median = stats::median(curator_minus_audience_log_odds),
    estimate_min = min(curator_minus_audience_log_odds),
    estimate_max = max(curator_minus_audience_log_odds),
    estimate_sd = if (.N > 1L) stats::sd(curator_minus_audience_log_odds) else 0,
    sign_changes_across_draws = min(curator_minus_audience_log_odds) < 0 &
      max(curator_minus_audience_log_odds) > 0
  ), by = term]
  sensitivity_summary[, feature := labels[term]]

  summary <- data.table(
    statistic = c(
      "candidate_scope", "development_articles", "paper2_test_articles",
      "development_candidate_comments", "paper2_test_candidate_comments",
      "development_curator_selections", "development_stacked_rows", "tie_draws_fitted",
      "article_selector_strata", "watermark"
    ),
    value = c(
      scope,
      uniqueN(development$story_id),
      uniqueN(test$story_id),
      nrow(development),
      nrow(test),
      sum(development$curator_selected),
      2L * nrow(development),
      tie_draws,
      2L * uniqueN(development$story_id),
      provenance$watermark
    )
  )

  scope_output <- file.path(output_root, scope)
  dir.create(scope_output, recursive = TRUE, showWarnings = FALSE)
  fwrite(scaling, file.path(scope_output, "feature_scaling.csv"))
  fwrite(coefficients, file.path(scope_output, "model_coefficients.csv"))
  fwrite(associations, file.path(scope_output, "selector_associations.csv"))
  fwrite(probabilities, file.path(scope_output, "probability_contrasts.csv"))
  fwrite(diagnostics, file.path(scope_output, "model_diagnostics.csv"))
  fwrite(sensitivity, file.path(scope_output, "tie_sensitivity_draws.csv"))
  fwrite(sensitivity_summary, file.path(scope_output, "tie_sensitivity_summary.csv"))
  fwrite(summary, file.path(scope_output, "sample_summary.csv"))
  test_scores <- score_frozen_model(primary, test, features)
  set(test_scores, j = "split_role", value = "paper2_test")
  set(test_scores, j = "score_source", value = "sealed_paper2_test")
  test_metrics <- evaluate_frozen_scores(test_scores)
  set(test_metrics, j = "split_role", value = "paper2_test")
  test_metric_summary <- bootstrap_metric_summary(
    test_metrics, test_bootstrap_draws, base_seed + if (scope == "root") 0L else 10000L
  )
  set(test_metric_summary, j = "split_role", value = "paper2_test")
  chance_metrics <- chance_ranking_metrics(test_scores)
  set(chance_metrics, j = "split_role", value = "paper2_test")
  chance_metric_summary <- bootstrap_metric_summary(
    chance_metrics,
    test_bootstrap_draws,
    base_seed + if (scope == "root") 20000L else 30000L
  )
  set(chance_metric_summary, j = "split_role", value = "paper2_test")
  above_chance_summary <- bootstrap_above_chance(
    test_metrics,
    chance_metrics,
    test_bootstrap_draws,
    base_seed + if (scope == "root") 40000L else 50000L
  )
  set(above_chance_summary, j = "split_role", value = "paper2_test")
  test_tie_sensitivity <- rbindlist(lapply(seq_len(tie_draws), function(draw) {
    draw_scores <- score_frozen_model(primary, test, features, audience_draw = draw)
    draw_metrics <- evaluate_frozen_scores(draw_scores)[selector == "audience"]
    set(draw_metrics, j = "audience_tie_draw", value = draw)
    draw_metrics
  }))
  set(test_tie_sensitivity, j = "split_role", value = "paper2_test")
  arrow::write_parquet(test_scores, file.path(scope_output, "test_scores_long.parquet"))
  test_scores_wide <- dcast(test_scores, story_id + comment_id ~ selector, value.var = "score")
  setnames(test_scores_wide, c("audience", "curator"),
    c("regression_audience_score", "regression_curator_score"))
  set(test_scores_wide, j = "split_role", value = "paper2_test")
  set(test_scores_wide, j = "score_source", value = "sealed_paper2_test")
  arrow::write_parquet(test_scores_wide, file.path(scope_output, "test_scores_wide.parquet"))
  arrow::write_parquet(test_metrics, file.path(scope_output, "test_article_metrics.parquet"))
  arrow::write_parquet(
    test_metric_summary, file.path(scope_output, "test_metric_summary.parquet")
  )
  arrow::write_parquet(
    chance_metrics, file.path(scope_output, "test_chance_article_metrics.parquet")
  )
  arrow::write_parquet(
    chance_metric_summary,
    file.path(scope_output, "test_chance_metric_summary.parquet")
  )
  arrow::write_parquet(
    above_chance_summary,
    file.path(scope_output, "test_above_chance_summary.parquet")
  )
  arrow::write_parquet(
    test_tie_sensitivity,
    file.path(scope_output, "test_tie_sensitivity_metrics.parquet")
  )
  arrow::write_parquet(article_split, file.path(scope_output, "article_split.parquet"))
  saveRDS(primary, file.path(scope_output, "stacked_selection_model.rds"))
  writeLines(
    c(
      paste("Candidate scope:", scope),
      paste("Feature watermark:", provenance$watermark),
      paste("Development articles:", uniqueN(development$story_id)),
      paste("FORUM held-out test articles:", uniqueN(test$story_id)),
      paste("Development candidate comments:", nrow(development)),
      paste("FORUM held-out candidate comments:", nrow(test)),
      paste("Tie draws fitted:", tie_draws),
      paste("Shared split:", split_path),
      paste("Fit cache:", normalizePath(fit_cache_root, mustWork = FALSE)),
      paste("Fit cache schema:", fit_cache_schema),
      "Feature scaling and coefficients: development articles only.",
      "FORUM scores: frozen coefficients applied once to sealed test articles.",
      paste("FORUM metric uncertainty:", test_bootstrap_draws,
        "article-bootstrap draws conditional on the fitted model."),
      "Chance baseline: exact expectation under an article-specific random",
      "permutation preserving n candidates and k selections; model-minus-chance",
      "uncertainty uses paired article bootstrap draws.",
      paste("FORUM audience tie-sensitivity draws:", tie_draws,
        "(fixed scores; labels varied without refitting)."),
      "Article controls: article-by-selector conditional-logit strata.",
      "Uncertainty: standard errors clustered by article.",
      "Audience outcome: top-k final relative-vote selection.",
      paste("Covariance condition number:",
        diagnostics[statistic == "covariance_condition_number", value]),
      "Probability contrasts: heuristic logistic translations; exact fixed-k",
      "counterfactual inclusion probabilities are deferred for possible future work.",
      "Interpretation: retrospective within-discussion associations, not causal effects.",
      if (length(attr(primary, "fit_warnings"))) {
        paste("Fit warnings:", paste(attr(primary, "fit_warnings"), collapse = " | "))
      } else {
        "Fit warnings: none"
      }
    ),
    file.path(scope_output, "diagnostics.txt")
  )
  progress_message(paste("Completed candidate scope:", scope))
  list(scope = scope, associations = associations, probabilities = probabilities, summary = summary)
}

results <- lapply(requested_scopes, run_scope)
progress_message("Analysis and artifact generation completed successfully")
```

# Compact results

```{r results, echo=FALSE}
for (result in results) {
  cat("\n##", if (result$scope == "root") "Root candidates" else "All comments", "\n\n")
  print(knitr::kable(result$summary, caption = "Sample accounting"))
  print(knitr::kable(
    result$associations[, .(
      feature,
      audience_log_odds,
      curator_log_odds,
      curator_minus_audience_log_odds,
      difference_conf_low,
      difference_conf_high,
      difference_q_value_bh
    )],
    digits = 4,
    caption = "Selector-specific associations on a common selection scale"
  ))
}
```

# Interpretation limits

The audience label is based on final collection-time vote totals and therefore
reflects exposure, position, social influence, and retrospective survival as
well as preference. Author-history vote variables are also collection-snapshot
reputation measures. Curators include journalists, community-management staff,
and volunteer guides. The models estimate associations only within discussions
where at least one usable sticky selection was observed. Association estimates
describe the development half; predictive metrics describe only the held-out
test half.
