---
title: "Legacy gap regression (02C-consistent, streamlined)"
output: html_document
date: "2026-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache.path = ".cache/knitr/02E_better-gap-measures/"
)
library(dplyr)
```

## Legacy data dependency

This analysis uses the pre-2025 `comment_data_general` CSV. It is retained as a
methodological reference for the Paper 1 gap-score port and is not part of the
2025 inference pipeline.

## Load data

```{r}
file <- "data/comment_data_general_091022_untilarticle5874_final_redacted.csv"
comment_data_general <- read.csv(file)[, -1]

comment_data_general$article <- as.factor(comment_data_general$article)
comment_data_general$pinned_f <- as.logical(comment_data_general$pinned_f)
comment_data_general$is_root_comment <- as.logical(comment_data_general$is_root_comment)

if (!("votes_rel" %in% names(comment_data_general))) {
  comment_data_general$votes_rel <- comment_data_general$votes_pos - comment_data_general$votes_neg
}

cat("Data loaded:", nrow(comment_data_general), "comments\\n")
cat("Articles:", dplyr::n_distinct(comment_data_general$article), "\\n")
cat("Pinned comments:", sum(comment_data_general$pinned_f, na.rm = TRUE), "\\n")
```

## Variable mapping (02C -> 02E aggregate)

```{r}
variable_mapping_02c_to_02e <- data.frame(
  var_02c = c(
    "hours_since_article",
    "num_sentences",
    "user_follower",
    "article_comments",
    "votes_pos_mean",
    "votes_neg_mean",
    "genre1",
    "(extension) n/a",
    "(extension) n/a",
    "(extension) n/a"
  ),
  var_02e = c(
    "hours_since_article_editors_mean",
    "num_sentences_editors_mean",
    "user_follower_editors_mean",
    "article_comments",
    "votes_pos_mean",
    "votes_neg_mean",
    "genre1",
    "n_picks",
    "follower_diff",
    "controversy"
  ),
  notes = c(
    "Mean over Editors' Picks in each discussion",
    "Mean over Editors' Picks in each discussion",
    "Mean over Editors' Picks in each discussion",
    "Discussion-level count (all comments)",
    "Discussion-level mean (all comments)",
    "Discussion-level mean (all comments)",
    "Same article-level genre factor as 02C",
    "Gap-specific extension (selection intensity)",
    "Gap-specific extension (editors vs top user picks)",
    "Gap-specific extension (root-comment downvote share)"
  ),
  stringsAsFactors = FALSE
)

print(variable_mapping_02c_to_02e)
```

## Build discussion-level modeling dataset

```{r build-model-data}
normalised_selection_rank_gap <- function(scores, selected) {
  if (length(scores) != length(selected)) {
    stop("scores and selected must have the same length")
  }

  keep <- !is.na(scores) & !is.na(selected)
  scores <- scores[keep]
  selected <- as.logical(selected[keep])

  n_candidates <- length(scores)
  n_selected <- sum(selected)

  if (n_candidates < 2 || n_selected < 1) return(NA_real_)

  # The ranking contains no information about selection quality when every
  # eligible candidate is selected: best, random, and worst sets coincide.
  if (n_selected >= n_candidates) return(NA_real_)

  # Midranks ensure tied vote totals do not inherit arbitrary row-order ranks.
  audience_ranks <- rank(-scores, ties.method = "average")
  mean_selected_rank <- mean(audience_ranks[selected])

  best_mean_rank <- (n_selected + 1) / 2
  worst_mean_rank <- n_candidates - (n_selected - 1) / 2

  (mean_selected_rank - best_mean_rank) /
    (worst_mean_rank - best_mean_rank)
}

build_gap_model_df_consistent <- function(comment_data_general) {
  by_article <- split(comment_data_general, comment_data_general$article)

  one_article <- function(df) {
    roots <- df[df$is_root_comment == TRUE, , drop = FALSE]
    if (nrow(roots) < 2) return(NULL)

    # Editors' Picks and the audience ranking must use the same eligible set.
    # Non-root pins are treated as data anomalies and excluded from this metric.
    editors <- roots[roots$pinned_f == TRUE, , drop = FALSE]
    if (nrow(editors) == 0) return(NULL)

    n_picks <- nrow(editors)
    gap_score <- normalised_selection_rank_gap(
      scores = roots$votes_rel,
      selected = roots$pinned_f
    )

    # Extension features
    controversy <- mean(roots$votes_neg / (roots$votes_pos + roots$votes_neg + 1), na.rm = TRUE)

    # A stable secondary key makes the separate follower-difference covariate
    # reproducible when relative-vote totals are tied at the top-k boundary.
    top_user <- roots %>%
      arrange(desc(votes_rel), comment_id) %>%
      slice_head(n = n_picks)

    follower_diff <- mean(editors$user_follower, na.rm = TRUE) - mean(top_user$user_follower, na.rm = TRUE)

    data.frame(
      article = as.character(df$article[1]),
      gap_score = gap_score,
      hours_since_article_editors_mean = mean(editors$hours_since_article, na.rm = TRUE),
      num_sentences_editors_mean = mean(editors$num_sentences, na.rm = TRUE),
      user_follower_editors_mean = mean(editors$user_follower, na.rm = TRUE),
      article_comments = nrow(df),
      votes_pos_mean = mean(df$votes_pos, na.rm = TRUE),
      votes_neg_mean = mean(df$votes_neg, na.rm = TRUE),
      genre1 = as.character(df$genre1[1]),
      n_picks = n_picks,
      follower_diff = follower_diff,
      controversy = controversy,
      stringsAsFactors = FALSE
    )
  }

  out <- bind_rows(lapply(by_article, one_article))
  out
}

gap_model_df_consistent <- build_gap_model_df_consistent(comment_data_general)

cat("N discussions with picks:", nrow(gap_model_df_consistent), "\\n")
cat("N with undefined gap because every eligible root was selected:",
    sum(is.na(gap_model_df_consistent$gap_score)), "\\n")
```

## Preprocessing (02C-style transforms)

```{r preprocess-model-data}
preprocess_gap_predictors_02c_style <- function(gap_model_df_consistent) {
  gap_model_df_consistent <- gap_model_df_consistent %>%
    filter(
      !is.na(gap_score),
      !is.na(hours_since_article_editors_mean),
      !is.na(num_sentences_editors_mean),
      !is.na(user_follower_editors_mean),
      !is.na(article_comments),
      !is.na(votes_pos_mean),
      !is.na(votes_neg_mean),
      !is.na(genre1),
      !is.na(n_picks),
      !is.na(follower_diff),
      !is.na(controversy)
    )

  gap_model_df_consistent$genre1 <- relevel(as.factor(gap_model_df_consistent$genre1), ref = "/inland")

  log_vars <- c(
    "hours_since_article_editors_mean",
    "user_follower_editors_mean",
    "article_comments",
    "votes_pos_mean",
    "votes_neg_mean",
    "n_picks"
  )

  for (v in log_vars) {
    gap_model_df_consistent[[paste0("log_", v)]] <- log1p(gap_model_df_consistent[[v]])
  }

  z_vars <- c(
    "log_hours_since_article_editors_mean",
    "num_sentences_editors_mean",
    "log_user_follower_editors_mean",
    "log_article_comments",
    "log_votes_pos_mean",
    "log_votes_neg_mean",
    "log_n_picks",
    "follower_diff",
    "controversy"
  )

  for (v in z_vars) {
    gap_model_df_consistent[[paste0("z_", v)]] <- as.numeric(scale(gap_model_df_consistent[[v]]))
  }

  # Epsilon clipping for beta family support in (0,1)
  eps <- 1e-6
  gap_model_df_consistent$gap_beta <- pmin(pmax(gap_model_df_consistent$gap_score, eps), 1 - eps)

  gap_model_df_consistent
}

gap_model_df_consistent <- preprocess_gap_predictors_02c_style(gap_model_df_consistent)

cat("Post-filter N:", nrow(gap_model_df_consistent), "\\n")
cat("Genre baseline:", levels(gap_model_df_consistent$genre1)[1], "\\n")
```

## Consistency and transform checks

```{r checks}
# Metric invariants for a ten-item ranking with three selected items.
toy_scores <- 10:1
stopifnot(all.equal(
  normalised_selection_rank_gap(toy_scores, seq_along(toy_scores) <= 3),
  0
))
stopifnot(all.equal(
  normalised_selection_rank_gap(toy_scores, seq_along(toy_scores) >= 8),
  1
))
stopifnot(is.na(
  normalised_selection_rank_gap(toy_scores, rep(TRUE, length(toy_scores)))
))

# The expected mean selected rank under random selection is (N + 1) / 2,
# which maps exactly to 0.5 under the corrected top-k/bottom-k normalisation.
n_toy <- length(toy_scores)
k_toy <- 3
random_expected_gap <- (
  (n_toy + 1) / 2 - (k_toy + 1) / 2
) / (n_toy - k_toy)
stopifnot(all.equal(random_expected_gap, 0.5))

# Tied scores receive equal midranks rather than row-order-dependent ranks.
tied_ranks <- rank(-c(10, 9, 9, 9, 8), ties.method = "average")
stopifnot(identical(tied_ranks[2:4], c(3, 3, 3)))
stopifnot(all.equal(
  normalised_selection_rank_gap(rep(1, 5), c(TRUE, FALSE, TRUE, FALSE, FALSE)),
  0.5
))

required_core <- c(
  "gap_score", "gap_beta",
  "hours_since_article_editors_mean", "num_sentences_editors_mean",
  "user_follower_editors_mean", "article_comments",
  "votes_pos_mean", "votes_neg_mean", "genre1"
)
missing_required <- setdiff(required_core, names(gap_model_df_consistent))
if (length(missing_required) > 0) {
  stop("Missing required columns: ", paste(missing_required, collapse = ", "))
}

required_transformed <- c(
  "z_log_hours_since_article_editors_mean",
  "z_num_sentences_editors_mean",
  "z_log_user_follower_editors_mean",
  "z_log_article_comments",
  "z_log_votes_pos_mean",
  "z_log_votes_neg_mean",
  "z_log_n_picks",
  "z_follower_diff",
  "z_controversy"
)

stopifnot(all(required_transformed %in% names(gap_model_df_consistent)))
finite_ok <- sapply(gap_model_df_consistent[, required_transformed], function(x) all(is.finite(x)))
if (!all(finite_ok)) {
  stop("Non-finite values in transformed predictors: ", paste(names(finite_ok)[!finite_ok], collapse = ", "))
}

if (levels(gap_model_df_consistent$genre1)[1] != "/inland") {
  warning("genre1 baseline is not '/inland'; check available levels")
}

if (any(gap_model_df_consistent$gap_score < 0 | gap_model_df_consistent$gap_score > 1)) {
  stop("Corrected gap score is outside [0, 1]")
}

cat("All consistency/transform checks passed.\\n")
```

## Collinearity diagnostics

```{r collinearity}
num_core <- gap_model_df_consistent %>%
  select(
    z_log_hours_since_article_editors_mean,
    z_num_sentences_editors_mean,
    z_log_user_follower_editors_mean,
    z_log_article_comments,
    z_log_votes_pos_mean,
    z_log_votes_neg_mean
  )

num_ext <- gap_model_df_consistent %>%
  select(
    z_log_hours_since_article_editors_mean,
    z_num_sentences_editors_mean,
    z_log_user_follower_editors_mean,
    z_log_article_comments,
    z_log_votes_pos_mean,
    z_log_votes_neg_mean,
    z_log_n_picks,
    z_follower_diff,
    z_controversy
  )

cat("Core predictor correlation matrix:\\n")
print(round(cor(num_core, use = "pairwise.complete.obs"), 3))

cat("\\nExtension predictor correlation matrix:\\n")
print(round(cor(num_ext, use = "pairwise.complete.obs"), 3))

compute_vif_like <- function(df_num) {
  vars <- names(df_num)
  out <- data.frame(variable = vars, vif_like = NA_real_)
  for (i in seq_along(vars)) {
    y <- vars[i]
    x <- vars[-i]
    f <- as.formula(paste(y, "~", paste(x, collapse = " + ")))
    r2 <- summary(lm(f, data = df_num))$r.squared
    out$vif_like[i] <- 1 / (1 - r2)
  }
  out
}

vif_like_ext <- compute_vif_like(num_ext)
print(vif_like_ext %>% arrange(desc(vif_like)))
```

## Beta regressions (glmmTMB)

```{r beta-regressions}
fit_gap_beta_models <- function(gap_model_df_consistent) {
  if (!requireNamespace("glmmTMB", quietly = TRUE)) return(NULL)

  core_formula <- gap_beta ~
    z_log_hours_since_article_editors_mean +
    z_num_sentences_editors_mean +
    z_log_user_follower_editors_mean +
    z_log_article_comments +
    z_log_votes_pos_mean +
    z_log_votes_neg_mean +
    genre1

  extension_formula <- gap_beta ~
    z_log_hours_since_article_editors_mean +
    z_num_sentences_editors_mean +
    z_log_user_follower_editors_mean +
    z_log_article_comments +
    z_log_votes_pos_mean +
    z_log_votes_neg_mean +
    genre1 +
    z_log_n_picks +
    z_follower_diff +
    z_controversy

  sensitivity_formula <- gap_beta ~
    z_log_hours_since_article_editors_mean +
    z_num_sentences_editors_mean +
    z_log_user_follower_editors_mean +
    z_log_article_comments +
    z_log_votes_pos_mean +
    z_log_votes_neg_mean +
    genre1 +
    z_follower_diff +
    z_controversy

  list(
    core = glmmTMB::glmmTMB(core_formula, data = gap_model_df_consistent, family = glmmTMB::beta_family(link = "logit")),
    extension = glmmTMB::glmmTMB(extension_formula, data = gap_model_df_consistent, family = glmmTMB::beta_family(link = "logit")),
    sensitivity = glmmTMB::glmmTMB(sensitivity_formula, data = gap_model_df_consistent, family = glmmTMB::beta_family(link = "logit"))
  )
}

gap_beta_models <- fit_gap_beta_models(gap_model_df_consistent)

if (is.null(gap_beta_models)) {
  cat("Package 'glmmTMB' is not installed; regression section skipped.\\n")
} else {
  cat("=== CORE MODEL ===\\n")
  print(summary(gap_beta_models$core))

  cat("\\n=== EXTENSION MODEL ===\\n")
  print(summary(gap_beta_models$extension))

  cat("\\n=== SENSITIVITY MODEL (NO n_picks) ===\\n")
  print(summary(gap_beta_models$sensitivity))

  model_comparison <- data.frame(
    model = c("core", "extension", "sensitivity_no_n_picks"),
    AIC = c(AIC(gap_beta_models$core), AIC(gap_beta_models$extension), AIC(gap_beta_models$sensitivity)),
    BIC = c(BIC(gap_beta_models$core), BIC(gap_beta_models$extension), BIC(gap_beta_models$sensitivity))
  )
  cat("\\n=== MODEL COMPARISON ===\\n")
  print(model_comparison)
}
```

## Export model data and coefficients

```{r export-model-data, include=FALSE}
export_gap_model_tables <- function(gap_model_df_consistent, model_list) {
  write.csv(gap_model_df_consistent, "data/gap_beta_model_data_consistent.csv", row.names = FALSE)
  write.csv(gap_model_df_consistent, "data/gap_beta_model_data.csv", row.names = FALSE)

  if (is.null(model_list)) return(invisible(NULL))

  coef_to_df <- function(m, model_name) {
    out <- as.data.frame(summary(m)$coefficients$cond)
    out$term <- rownames(out)
    rownames(out) <- NULL
    out <- out %>%
      select(term, Estimate, `Std. Error`, `z value`, `Pr(>|z|)`) %>%
      mutate(model = model_name)
    out
  }

  beta_coef_core <- coef_to_df(model_list$core, "core")
  beta_coef_extension <- coef_to_df(model_list$extension, "extension")
  beta_coef_sensitivity <- coef_to_df(model_list$sensitivity, "sensitivity")

  write.csv(beta_coef_core, "data/gap_beta_model_coefficients_core.csv", row.names = FALSE)
  write.csv(beta_coef_extension, "data/gap_beta_model_coefficients_extension.csv", row.names = FALSE)
  write.csv(beta_coef_sensitivity, "data/gap_beta_model_coefficients_sensitivity.csv", row.names = FALSE)
  write.csv(bind_rows(beta_coef_core, beta_coef_extension, beta_coef_sensitivity),
            "data/gap_beta_model_coefficients.csv", row.names = FALSE)

  invisible(list(core = beta_coef_core, extension = beta_coef_extension, sensitivity = beta_coef_sensitivity))
}

coef_exports <- export_gap_model_tables(gap_model_df_consistent, gap_beta_models)
```
